---
title: "The property tax dividend from Edmonton's RS Zone infill"
date: "2026-07-12"
author: "Jacob Dawang"
categories: [housing, edmonton, zoning]
description: "Redeveloping an old house into a 5-8 unit rowhome, 2-4 unit plex, or backyard home doesn't just add homes, it multiplies the property's assessed value. Here's what all of it is worth to Edmonton's property tax base since the Zoning Bylaw Renewal took effect."
link-external-newwindow: true
link-external-icon: false
execute:
  cache: false
draft: false
---

Edmonton's Zoning Bylaw Renewal (ZBR) has spent two and a half years adding new homes to existing neighbourhoods, mostly through three kinds of small-scale infill the RS Zone allows: the extremely successful 5 to 8 unit rowhome, the smaller 2 to 4 unit plex, and the backyard home (garden suite) added behind a retained house. [In an ongoing series of blog posts](../q2-building-permits/index.qmd), I track how the RS zone is slowly transforming our city. This post asks a different question: what is all of that redevelopment worth to the City's property tax base?

When a single-detached home in a mature neighbourhood becomes a 5 to 8 unit rowhome or a 2 to 4 unit plex, or gains a backyard home, its assessed value increases, adding to the City's property tax revenue.

Using building permits from 2024, the first year the RS Zone was in force, matched against the City's property assessment data, I measure that uplift directly for completed projects in each category, then fit a single pooled regression model across all three to estimate what it adds up to across the full 2024-2026 pipeline. The 5 to 8 unit rowhome is the largest single category by far, so it gets the detailed worked example; the same method is then applied to the other two.

## Key numbers

-   **$1.73 billion**: The amount that small-scale infill — 5 to 8 unit rowhomes, 2 to 4 unit plexes, and backyard homes — in the RS Zone from 2024-2026 will add to the property tax base at steady state.

-   **$14 million per year**: The amount of incremental property tax that the City of Edmonton will collect from RS Zone small-scale infill at steady state.

## Data {#sec-data}

This analysis combines three of Edmonton's Open Data sources. Edmonton's [general building permit data](https://data.edmonton.ca/Urban-Planning-Economy/General-Building-Permits/24uj-dj8v/about_data), the same source used in the [Q2 2026 building permit report](../q2-building-permits/index.qmd), identifies which properties were redeveloped into 5 to 8 unit rowhomes, 2 to 4 unit plexes, or backyard homes, and when. Edmonton's [current calendar year property assessment data](https://data.edmonton.ca/City-Administration/Property-Assessment-Data-Current-Calendar-Year-/q7d6-ambg/about_data) gives each property's assessed value as of 2026, and Edmonton's [historical property assessment data](https://data.edmonton.ca/City-Administration/Property-Assessment-Data-Historical-/qi6a-xuwt/about_data) gives assessed values back to 2012, including 2024 and 2025.

The unit of analysis throughout this post is the **lot**, not the permit — a redevelopment can take more than one permit, and a permit alone can double-count or miss a lot's true uplift. Three situations come up:

1. **One permit on a lot.** The common case: match the permit to that lot's one pre- and post-redevelopment assessment.
2. **More than one permit on the same lot.** A fourplex redevelopment plus a separate garden-suite permit on the same never-subdivided lot, for instance — sometimes issued a year or more apart. These are one redevelopment, not two, so I add the permits together: their units sum, and the one shared assessed value covers both (a garden suite is almost always assessed as part of its host property, not as a separate account). Categorizing by the combined unit count, not any single permit's own count, means a fourplex-plus-suite lot with 6 total homes counts as part of the 5 to 8 unit rowhome story, not the 2 to 4 unit plex one.
3. **A lot split into two or more child lots, each separately permitted** — a duplex-style subdivision under a brand-new plan, for instance. Each child gets its own completed-building assessment, but only one of them, if any, can be traced back to the original, pre-split lot's baseline value. I assign an equal share of that original value to each child as its baseline.

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(sf)
library(gt)
library(scales)
library(mountainmathHelpers)
library(jdawangHelpers)
library(mgcv)

options(
  cancensus.api_key = Sys.getenv("CM_API_KEY"),
  cancensus.cache_path = Sys.getenv("CM_CACHE_PATH"),
  tongfen.cache_path = Sys.getenv("TONGFEN_CACHE_PATH"),
  nextzen_API_key = Sys.getenv("NEXTZEN_API_KEY")
)

label_dollars_adaptive <- function(x) {
  if (abs(x) >= 1e9) {
    label_dollar(scale = 1e-9, suffix = "B", accuracy = 0.01)(x)
  } else if (abs(x) >= 1e8) {
    label_dollar(scale = 1e-6, suffix = "M", accuracy = 1)(x)
  } else {
    label_dollar(scale = 1e-6, suffix = "M", accuracy = 0.1)(x)
  }
}

# Parses a permit's legal description into a (plan, block, lot) key, used to
# match permits to assessments (@sec-matching below).
parse_permit_legal <- function(x) {
  m <- str_match(x, "Plan\\s+(\\S+)\\s+Blk\\s+(\\S+)\\s+Lots?\\s+(\\S+)")
  tibble(plan = m[, 2], block = m[, 3], lot = m[, 4])
}
normalize_key <- function(plan, block, lot) {
  # A missing component must yield a real NA, not the literal string "NA|NA|NA":
  # otherwise every parcel with an incomplete legal description (e.g. a
  # letter-lot "Plan X Lot H" with no block) collapses onto one shared key and
  # gets falsely matched to a single arbitrary assessment row.
  if_else(
    is.na(plan) | is.na(block) | is.na(lot),
    NA_character_,
    paste(
      toupper(str_trim(plan)),
      toupper(str_trim(block)),
      toupper(str_trim(lot)),
      sep = "|"
    )
  )
}

# Strips the trailing letter off a legal-description lot number (e.g. "40A"
# -> "40"), grouping the two (or more) child lots a parcel was subdivided
# into under one new plan. Used to share one pre-subdivision baseline across
# a lot's children (@sec-consolidating, situation 3). Legal descriptions
# that don't match the "Plan X Blk Y Lot Z" format at all (e.g. "Condo
# Common Area (Plan X)") fall back to the raw string as their own key, so
# every one of them is its own singleton family -- otherwise they'd all
# collapse onto the same "NA|NA|NA" key and get treated as siblings of each
# other, sharing a nonsense baseline across completely unrelated lots.
parent_lot_key <- function(legal_description) {
  parsed <- parse_permit_legal(legal_description)
  lot_numeric <- str_extract(parsed$lot, "^[0-9]+")
  key <- normalize_key(parsed$plan, parsed$block, lot_numeric)
  if_else(
    is.na(parsed$plan) | is.na(parsed$block) | is.na(lot_numeric),
    legal_description,
    key
  )
}

# Parses a permit's civic address into a (house_number, street_name) key,
# used as a fallback (@sec-matching) for permits with no geometry and a
# legal description on a subdivision plan too new to appear in either
# assessment roll. Strips an optional leading "UNIT, "/"BSMT, " prefix and
# any letter suffix on the house number itself (e.g. the "G" that marks a
# garden-suite building at an existing address), since both assessment
# rolls' house_number field is purely numeric.
parse_permit_address <- function(x) {
  x_no_unit <- str_replace(x, "^[^,]+,\\s*", "")
  m <- str_match(x_no_unit, "^(\\d+)[A-Za-z]?\\s*-\\s*(.+)$")
  tibble(house_number = m[, 2], street_name = toupper(str_trim(m[, 3])))
}

# The 2026 assessment only reliably reflects buildings finished by about
# this date (see @sec-assessment-catch-up); used as the completion cutoff
# for every category's measured cohort.
occupancy_cutoff <- as.Date("2025-09-30")

bp_raw <- get_edmonton_building_permit_data(
  cache_dir = Sys.getenv("EDMONTON_BP_CACHE_PATH"),
  end_date = "2026-06-30"
)
crs <- lambert_conformal_conic_at(bp_raw)

bp <- bp_raw |>
  st_transform(crs) |>
  mutate(permit_row_id = row_number()) |>
  filter(year < 2026 | (year == 2026 & month_number <= 6)) |>
  filter_edmonton_residential() |>
  filter(units_added >= 1, work_type != "(04) Excavation") |>
  add_edmonton_project_type() |>
  add_edmonton_suite_info() |>
  add_edmonton_neighbourhood_type() |>
  # Some permits carry an occupancy_granted_date that predates their own
  # issue date -- a known data error (e.g. a 2024 new-build stamped with a
  # 2022/2023 occupancy). Blank those dates so the bad date can't put a lot
  # into the measured cohort's occupancy cutoff. The lot itself is not
  # dropped: with no trustworthy completion date it simply falls to the
  # model-predicted side of the pipeline (all_sites_pred doesn't filter on
  # occupancy), so it's still counted in the steady-state uplift.
  mutate(
    occupancy_granted_date = if_else(
      !is.na(date_issued) & occupancy_granted_date < date_issued,
      as.Date(NA),
      occupancy_granted_date
    )
  )

# LRT stops (existing + under construction), same setup as the Q2 2026 report
future_stops <- read_sf("../../../data/future_lrt_stops.geojson") |>
  filter(
    !(stop_name %in%
      c(
        "Castle Downs Station",
        "145 Ave Station",
        "137 Ave Station",
        "132 Ave Station"
      ))
  )

transit_stops <- load_edmonton_transit_stops(
  gtfs_path = "../../../data/ca-alberta-edmonton-transit-system-gtfs-714.zip",
  service_date = as.Date("2023-11-09"),
  future_stops = future_stops,
  crs = crs
)

# The full universe this post is about: every mature-neighbourhood RS Zone
# permit that could be part of a 5-8 unit rowhome, 2-4 unit plex, or
# backyard-home redevelopment, permitted since ZBR took effect. Categories
# aren't assigned here -- that only happens once same-lot permits are
# consolidated (@sec-consolidating), since a permit's own unit count can be
# misleading (a lot's *combined* units are what matters).
redevelopment_permits <- bp |>
  filter(
    zoning == "RS",
    work_type == "(01) Building - New",
    neighbourhood_type == "Mature",
    between(units_added, 2, 8) | project_type == "Backyard House"
  ) |>
  add_transit_distance(transit_stops)
```

```{r}
#| label: fetch-assessment-data
# This is a one-off fetch specific to this post: Edmonton's property
# assessment data hasn't been used on this blog before, so this helper
# isn't (yet) part of jdawangHelpers.
fetch_socrata_csv <- function(resource, query, cache_file) {
  cache_dir <- Sys.getenv("EDMONTON_BP_CACHE_PATH")
  path <- file.path(cache_dir, cache_file)
  if (!file.exists(path)) {
    url <- paste0(
      "https://data.edmonton.ca/resource/",
      resource,
      ".csv?",
      URLencode(query)
    )
    # These pulls are hundreds of MB; the default 60s timeout is far too
    # short. Download to a temp file and only promote it to the cache path
    # on success, so an interrupted download can't leave a truncated file
    # that later runs would silently reuse.
    old_timeout <- getOption("timeout")
    options(timeout = max(1800, old_timeout))
    on.exit(options(timeout = old_timeout), add = TRUE)
    tmp <- paste0(path, ".part")
    download.file(url, tmp, mode = "wb")
    file.rename(tmp, path)
  }
  read_csv(path, show_col_types = FALSE, col_types = cols(.default = "c"))
}

assessment_hist <- fetch_socrata_csv(
  "qi6a-xuwt",
  "$where=assessment_year in (2023,2024,2025)&$limit=2000000",
  "assessment_2023_2025.csv"
) |>
  mutate(
    assessment_year = as.integer(assessment_year),
    assessed_value = as.numeric(assessed_value),
    latitude = as.numeric(latitude),
    longitude = as.numeric(longitude)
  )

assessment_2026 <- fetch_socrata_csv(
  "q7d6-ambg",
  "$limit=600000",
  "assessment_2026.csv"
) |>
  mutate(
    assessed_value = as.numeric(assessed_value),
    latitude = as.numeric(latitude),
    longitude = as.numeric(longitude)
  )

tax_rates <- fetch_socrata_csv("pwis-wc4c", "$limit=10000", "tax_rates.csv") |>
  mutate(
    tax_year = as.integer(tax_year),
    amount_per_dollar_of_assessed_value = as.numeric(
      amount_per_dollar_of_assessed_value
    )
  )

assessment_2023 <- assessment_hist |> filter(assessment_year == 2023)
assessment_2024 <- assessment_hist |> filter(assessment_year == 2024)
assessment_2025 <- assessment_hist |> filter(assessment_year == 2025)

municipal_rate <- function(class, yr) {
  tax_rates |>
    filter(
      tax_year == yr,
      tax_rate_type == "Municipal",
      assessment_class == class
    ) |>
    pull(amount_per_dollar_of_assessed_value)
}
```

## Matching permits to assessments and consolidating lots {#sec-matching}

The key challenge is linking a specific building permit to the same physical property, before and after redevelopment.

### Matching a permit to its assessment record

Account numbers usually persist through redevelopment, but addresses often don't. Legal description (plan/block/lot), seems to be more stable across redevelopment. However, I have two fallbacks for the properties that don't match that way.

I parse each side's legal description into a `(plan, block, lot)` key and join on it. A building permit's legal description looks like `Plan 715HW Blk 25 Lot 22`; the historical assessment data records the same parcel as `Plan: 715HW  Block: 25  Lot: 22`.

```{r}
#| label: legal-key-matching
parse_assessment_legal <- function(x) {
  m <- str_match(x, "Plan:\\s*(\\S+)\\s+Block:\\s*(\\S+)\\s+Lot:\\s*(\\S+)")
  tibble(plan = m[, 2], block = m[, 3], lot = m[, 4])
}

assessment_2024_keyed <- assessment_2024 |>
  bind_cols(parse_assessment_legal(assessment_2024$legal_description)) |>
  mutate(key = normalize_key(plan, block, lot)) |>
  filter(!is.na(key)) |>
  distinct(key, .keep_all = TRUE)

# Reusable matcher: join a set of permits to their 2024 baseline assessed
# value by legal-description key. Used for the full permit universe here,
# and again later for the model's own baseline lookups.
match_baseline_2024 <- function(cohort_sf) {
  cohort_sf |>
    st_drop_geometry() |>
    bind_cols(parse_permit_legal(cohort_sf$legal_description)) |>
    mutate(key = normalize_key(plan, block, lot)) |>
    left_join(
      assessment_2024_keyed |>
        select(key, account_number, value_2024 = assessed_value),
      by = "key"
    )
}

baseline_matched <- match_baseline_2024(redevelopment_permits)
n_matched_legal <- sum(!is.na(baseline_matched$account_number))
```

That legal-key match links most properties to their 2024 baseline assessed value directly. The first fallback is geographic: the nearest 2024 assessment point within 25m, for lots consolidated from two legal titles into one, or where the legal description format doesn't parse identically on both sides.

```{r}
#| label: geo-fallback-2024
# Reusable: fall back to the nearest 2024 assessment point within 25m for
# any permit the legal-key match missed.
geo_fallback_baseline <- function(matched_df, cohort_sf) {
  still_unmatched <- matched_df |> filter(is.na(account_number))
  if (nrow(still_unmatched) == 0) {
    return(matched_df)
  }

  unmatched_sf <- cohort_sf |>
    filter(permit_row_id %in% still_unmatched$permit_row_id)
  assessment_2024_sf <- assessment_2024 |>
    filter(!is.na(latitude), !is.na(longitude)) |>
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
    st_transform(crs)

  nearest_idx <- st_nearest_feature(unmatched_sf, assessment_2024_sf)
  dist_m <- as.numeric(st_distance(
    unmatched_sf,
    assessment_2024_sf[nearest_idx, ],
    by_element = TRUE
  ))

  geo_matches <- tibble(
    permit_row_id = unmatched_sf$permit_row_id,
    geo_account = assessment_2024_sf$account_number[nearest_idx],
    geo_value_2024 = assessment_2024_sf$assessed_value[nearest_idx],
    geo_dist_m = dist_m
  ) |>
    filter(geo_dist_m <= 25)

  matched_df |>
    left_join(geo_matches, by = "permit_row_id") |>
    mutate(
      account_number = coalesce(account_number, geo_account),
      value_2024 = coalesce(value_2024, geo_value_2024)
    ) |>
    select(-any_of(c("geo_account", "geo_value_2024", "geo_dist_m")))
}

baseline_matched <- geo_fallback_baseline(
  baseline_matched,
  redevelopment_permits
)
n_matched_after_geo <- sum(!is.na(baseline_matched$account_number))
```

The 25m fallback closes most of the remaining gap. What's left mostly carries a legal description reissued under a brand-new subdivision plan, with no permit coordinates for the 25m fallback to use either.

The civic address usually still resolves even when the legal description doesn't: Edmonton's assessment rolls index every parcel by house number and street name independently of its legal description.

```{r}
#| label: address-fallback-baseline
# Third-tier fallback: for permits with no geometry (so the 25m fallback
# above has nothing to search from) whose legal description is on a
# subdivision plan too new for the 2024 roll, the civic address usually
# still resolves. Restricted to non-suite assessment rows, since a
# condo/suite shares its civic address with many unrelated units at very
# different values.
address_fallback_baseline <- function(matched_df, cohort_sf) {
  still_unmatched <- matched_df |> filter(is.na(account_number))
  if (nrow(still_unmatched) == 0) {
    return(matched_df)
  }

  addr_cohort <- cohort_sf |>
    st_drop_geometry() |>
    filter(permit_row_id %in% still_unmatched$permit_row_id) |>
    select(permit_row_id, address)
  addr_parsed <- addr_cohort |>
    bind_cols(parse_permit_address(addr_cohort$address)) |>
    mutate(house_number = as.character(as.numeric(house_number)))

  assessment_2024_addr <- assessment_2024 |>
    filter(is.na(suite)) |>
    mutate(
      house_number = as.character(as.numeric(house_number)),
      street_name = toupper(str_trim(street_name))
    ) |>
    distinct(house_number, street_name, .keep_all = TRUE)

  addr_matches <- addr_parsed |>
    left_join(
      assessment_2024_addr |>
        select(
          house_number,
          street_name,
          addr_account = account_number,
          addr_value_2024 = assessed_value
        ),
      by = c("house_number", "street_name")
    ) |>
    filter(!is.na(addr_account)) |>
    select(permit_row_id, addr_account, addr_value_2024)

  matched_df |>
    left_join(addr_matches, by = "permit_row_id") |>
    mutate(
      account_number = coalesce(account_number, addr_account),
      value_2024 = coalesce(value_2024, addr_value_2024)
    ) |>
    select(-any_of(c("addr_account", "addr_value_2024")))
}

baseline_matched <- address_fallback_baseline(
  baseline_matched,
  redevelopment_permits
)
n_matched_baseline <- sum(!is.na(baseline_matched$account_number))
n_total_permits <- nrow(baseline_matched)
```

Combining all three tiers links **`r n_matched_baseline`** of `r n_total_permits` permits (`r percent(n_matched_baseline / n_total_permits, accuracy = 1)`) to a 2024 baseline assessed value directly. Some of the rest still turn out to have a baseline after all: @sec-consolidating covers how.

In most cases the account number persists through redevelopment, so I use it to look up the completed building's 2026 assessed value, with the same two fallbacks (25m geographic proximity, then civic address) for the properties issued a brand new account number when redeveloped.

```{r}
#| label: join-2026
# Reusable: join a baseline-matched permit to its 2026 outcome value by
# account_number, falling back to (1) the nearest 2026 assessment point
# within 25m, then (2) the permit's own post-redevelopment civic address,
# for the permits geometry can't reach. Unlike the baseline matcher, this
# must NOT require account_number up front: a permit can lack any 2024
# baseline (its lot was split into a brand-new address with no
# pre-existing account) yet still have a perfectly good completed 2026
# value once its own address resolves -- so every permit stays in play
# here, whether or not the baseline matchers above found it anything.
match_outcome_2026 <- function(matched_df, cohort_sf) {
  joined <- matched_df |>
    left_join(
      assessment_2026 |>
        select(account_number, value_2026 = assessed_value, mill_class_1),
      by = "account_number"
    )

  missing_2026 <- joined |> filter(is.na(value_2026))
  if (nrow(missing_2026) > 0) {
    missing_2026_sf <- cohort_sf |>
      filter(permit_row_id %in% missing_2026$permit_row_id)
    assessment_2026_sf <- assessment_2026 |>
      filter(!is.na(latitude), !is.na(longitude)) |>
      st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
      st_transform(crs)

    nearest_idx <- st_nearest_feature(missing_2026_sf, assessment_2026_sf)
    geo_2026 <- tibble(
      permit_row_id = missing_2026_sf$permit_row_id,
      geo_value_2026 = assessment_2026_sf$assessed_value[nearest_idx],
      geo_mill_class = assessment_2026_sf$mill_class_1[nearest_idx],
      geo_dist_m = as.numeric(st_distance(
        missing_2026_sf,
        assessment_2026_sf[nearest_idx, ],
        by_element = TRUE
      ))
    ) |>
      filter(geo_dist_m <= 25)

    joined <- joined |>
      left_join(geo_2026, by = "permit_row_id") |>
      mutate(
        value_2026 = coalesce(value_2026, geo_value_2026),
        mill_class_1 = coalesce(mill_class_1, geo_mill_class)
      ) |>
      select(-any_of(c("geo_value_2026", "geo_mill_class", "geo_dist_m")))
  }

  still_missing_2026 <- joined |> filter(is.na(value_2026))
  if (nrow(still_missing_2026) == 0) {
    return(joined)
  }

  addr_cohort <- cohort_sf |>
    st_drop_geometry() |>
    filter(permit_row_id %in% still_missing_2026$permit_row_id) |>
    select(permit_row_id, address)
  addr_parsed <- addr_cohort |>
    bind_cols(parse_permit_address(addr_cohort$address)) |>
    mutate(house_number = as.character(as.numeric(house_number)))

  assessment_2026_addr <- assessment_2026 |>
    filter(is.na(suite)) |>
    mutate(
      house_number = as.character(as.numeric(house_number)),
      street_name = toupper(str_trim(street_name))
    ) |>
    distinct(house_number, street_name, .keep_all = TRUE)

  addr_matches_2026 <- addr_parsed |>
    left_join(
      assessment_2026_addr |>
        select(
          house_number,
          street_name,
          addr_value_2026 = assessed_value,
          addr_mill_class = mill_class_1
        ),
      by = c("house_number", "street_name")
    ) |>
    filter(!is.na(addr_value_2026)) |>
    select(permit_row_id, addr_value_2026, addr_mill_class)

  joined |>
    left_join(addr_matches_2026, by = "permit_row_id") |>
    mutate(
      value_2026 = coalesce(value_2026, addr_value_2026),
      mill_class_1 = coalesce(mill_class_1, addr_mill_class)
    ) |>
    select(-any_of(c("addr_value_2026", "addr_mill_class")))
}

matched_outcome <- match_outcome_2026(baseline_matched, redevelopment_permits)
n_matched_outcome <- sum(!is.na(matched_outcome$value_2026))
```

That gives **`r n_matched_outcome`** of `r n_total_permits` permits (`r percent(n_matched_outcome / n_total_permits, accuracy = 1)`) a completed 2026 assessed value, whether or not that same permit matched a 2024 baseline directly.

### Consolidating same-lot permits and lot splits {#sec-consolidating}

Every permit above is matched on its own, but the unit of analysis is the lot. I now collapse same-lot permits together (situation 2 from the @sec-data) and share a baseline across split lots (situation 3), across the whole 2024-2026 timeframe. For example, a fourplex permitted in 2024 and its garden suite permitted in 2025 are still one lot. Treating them as two separate, single-year events would let the same original assessed value get counted twice.

```{r}
#| label: build-sites
first_non_na <- function(x) {
  v <- x[!is.na(x)]
  if (length(v) == 0) {
    return(x[1])
  }
  v[1]
}

# A lot's completed value can come from more than one permit's own matched
# account (e.g. a fourplex and its later garden suite occasionally resolve
# to two different final accounts). Summing over *distinct* (value,
# assessment class) pairs, rather than over every constituent permit,
# correctly handles both a lot that resolves to one shared account (the
# distinct set has one member, so nothing is double-counted) and a lot
# whose permits genuinely each earned their own separate completed value.
sum_distinct_value_2026 <- function(value_2026, mill_class_1) {
  d <- tibble(value_2026 = value_2026, mill_class_1 = mill_class_1) |>
    filter(!is.na(value_2026)) |>
    distinct()
  if (nrow(d) == 0) {
    return(NA_real_)
  }
  sum(d$value_2026)
}

# Situations 1 & 2: collapse every permit sharing an exact legal
# description -- regardless of permit year -- into one row per lot.
# units_added sums (each permit represents real added units);
# occupancy_granted_date takes the latest (the lot isn't "done" until its
# last permit is); the "primary" permit -- the one with the most units, a
# lot's main redevelopment rather than an added suite -- supplies the
# site's year, address, neighbourhood, and transit distance.
build_sites <- function(matched) {
  m <- matched |> st_drop_geometry()

  primary <- m |>
    arrange(legal_description, desc(units_added), year) |>
    distinct(legal_description, .keep_all = TRUE) |>
    select(legal_description, year, address, neighbourhood, distance_from_lrt)

  totals <- m |>
    group_by(legal_description) |>
    summarise(
      units_added = sum(units_added),
      occupancy_granted_date = if (all(is.na(occupancy_granted_date))) {
        as.Date(NA)
      } else {
        max(occupancy_granted_date, na.rm = TRUE)
      },
      account_number = first_non_na(account_number),
      value_2024 = first_non_na(value_2024),
      value_2026 = sum_distinct_value_2026(value_2026, mill_class_1),
      mill_class_1 = first_non_na(mill_class_1),
      all_backyard = all(project_type == "Backyard House"),
      .groups = "drop"
    )

  primary |> left_join(totals, by = "legal_description")
}

# Situation 3: share one baseline across a split lot's children. A family
# (same plan/block, lot letter stripped) only gets its baseline shared if
# exactly one distinct baseline value is found anywhere among its children
# -- that's the signature of a lot that split after the 2024 snapshot, with
# only one child (or none) resolving the original parcel's value directly.
# A family where two children already carry two genuinely different
# baseline values is left alone: that's an old, already-separate pair of
# lots that happen to share a lot-letter naming pattern, not a recent
# split, and sharing would silently discard one of their real baselines.
share_split_baselines <- function(sites) {
  keyed <- sites |> mutate(parent_key = parent_lot_key(legal_description))

  family_baseline <- keyed |>
    filter(!is.na(value_2024)) |>
    distinct(parent_key, value_2024) |>
    add_count(parent_key, name = "n_distinct_baselines") |>
    filter(n_distinct_baselines == 1) |>
    select(parent_key, family_value_2024 = value_2024)

  family_size <- keyed |> count(parent_key, name = "n_children")

  keyed |>
    left_join(family_baseline, by = "parent_key") |>
    left_join(family_size, by = "parent_key") |>
    mutate(
      shared_baseline = !is.na(family_value_2024) & n_children > 1,
      value_2024 = if_else(
        shared_baseline,
        family_value_2024 / n_children,
        value_2024
      )
    ) |>
    select(-family_value_2024, -n_children, -parent_key)
}

sites <- build_sites(matched_outcome) |>
  share_split_baselines() |>
  mutate(
    category = case_when(
      all_backyard ~ "backyard",
      units_added <= 4 ~ "plex",
      TRUE ~ "rowhome"
    ),
    # A site with no geometry (the same missing-coordinates problem the
    # address fallback above works around) has no LRT distance either.
    # Falling back to the citywide median here, once, means every
    # downstream consumer of `sites` already has a usable covariate rather
    # than needing its own fallback.
    distance_from_lrt = coalesce(
      distance_from_lrt,
      median(distance_from_lrt, na.rm = TRUE)
    )
  )

n_sites <- nrow(sites)
n_sites_matched <- sum(!is.na(sites$value_2024) & !is.na(sites$value_2026))
n_shared_baseline <- sum(sites$shared_baseline)

stopifnot(
  # Every site is a distinct lot by construction.
  n_distinct(sites$legal_description) == n_sites,
  # Consolidating and sharing baselines moves value around but never
  # drops or duplicates a permit's units.
  sum(sites$units_added) == sum(matched_outcome$units_added)
)
```

Consolidating collapses `r n_total_permits` permits into **`r n_sites`** distinct lots. Sharing a baseline across split lots (situation 3) resolves `r n_shared_baseline` of them that the permit-level matchers above couldn't reach on their own. Altogether, **`r n_sites_matched`** lots (`r percent(n_sites_matched / n_sites, accuracy = 1)`) end up with both a 2024 baseline and a 2026 completed value.

```{r}
#| label: category-match-rates
category_match_rates <- sites |>
  group_by(category) |>
  summarise(
    n = n(),
    n_matched = sum(!is.na(value_2024) & !is.na(value_2026)),
    .groups = "drop"
  ) |>
  mutate(pct = percent(n_matched / n, accuracy = 1))

rowhome_match_rate <- category_match_rates |>
  filter(category == "rowhome") |>
  pull(pct)
plex_match_rate <- category_match_rates |>
  filter(category == "plex") |>
  pull(pct)
backyard_match_rate <- category_match_rates |>
  filter(category == "backyard") |>
  pull(pct)
```

The match rate holds up close to evenly across all three categories: `r rowhome_match_rate` of 5 to 8 unit rowhome lots, `r plex_match_rate` of 2 to 4 unit plex lots, and `r backyard_match_rate` of backyard-home lots resolve both a baseline and a completed value. Plexes are the category most likely to involve a lot split (a duplex-style subdivision into two semi-detached halves is a common way to build one), which is exactly the situation the split-baseline sharing above is for.

## Measuring the uplift: the 2024 cohort

I measure completed 2024 lots in all three categories — 5 to 8 unit rowhomes, 2 to 4 unit plexes, and backyard homes — against the same assessment-lag and counterfactual-growth machinery. The 5 to 8 unit rowhome is by far the largest category, so I walk through that category first as an example, then apply the same method to the other two.

### Worked example: the 5 to 8 unit rowhome cohort

The 2026 property assessment is supposed to reflect each property's physical condition as of December 31, 2025. In principle, that means any 2024-permitted lot whose building had received occupancy by the end of 2025 should be assessed as completed in the 2026 assessment roll. In practice, the assessments lag, as @sec-assessment-catch-up checks directly.

```{r}
#| label: measurement-cohort
rowhome_2024 <- sites |>
  filter(category == "rowhome", year == 2024)

n_cohort_2024_all <- nrow(rowhome_2024)
n_occupancy_by_dec <- rowhome_2024 |>
  filter(
    !is.na(occupancy_granted_date),
    occupancy_granted_date <= as.Date("2025-12-31")
  ) |>
  nrow()

rowhome_2024_matched <- rowhome_2024 |>
  filter(!is.na(value_2024), !is.na(value_2026)) |>
  left_join(
    assessment_2025 |> select(account_number, value_2025 = assessed_value),
    by = "account_number"
  ) |>
  mutate(
    bucket = if_else(units_added <= 6, "5-6 units", "7+ units"),
    gross_uplift = value_2026 - value_2024
  )
n_rowhome_2024_matched <- nrow(rowhome_2024_matched)
```

Edmonton issued **`r n_cohort_2024_all`** mature-neighbourhood RS Zone 5-8 unit new-build lots in 2024, of which **`r n_occupancy_by_dec`** had received occupancy by the end of 2025. **`r n_rowhome_2024_matched`** of the `r n_cohort_2024_all` (`r percent(n_rowhome_2024_matched / n_cohort_2024_all, accuracy = 1)`) match both a 2024 baseline and a 2026 assessed value; I use the occupancy dates to decide which of those the 2026 assessment actually reflects as complete.

#### When does the assessment catch up? {#sec-assessment-catch-up}

Before treating every occupancy-granted building as "complete" in the 2026 data, it's worth checking whether the assessments actually keep up with late-year completions. @fig-assessment-lag plots each matched lot's gross uplift against its occupancy date. Buildings that received occupancy by roughly the end of September 2025 tend to show the full 5 to 8 unit rowhome uplift. For many October, November, and especially December completions, the uplift is much smaller and doesn't follow the same pattern — probably because their assessments still reflect an unfinished state.

```{r}
#| label: fig-assessment-lag
#| fig-cap: "5 to 8 unit rowhomes: gross assessed value uplift by occupancy date, showing the 2026 assessment only reliably reflects buildings finished by about September 2025"
rowhome_2024_matched |>
  filter(
    !is.na(occupancy_granted_date),
    occupancy_granted_date <= as.Date("2025-12-31")
  ) |>
  mutate(
    reflected = if_else(
      occupancy_granted_date <= occupancy_cutoff,
      "Occupancy by Sept 2025",
      "Occupancy Oct-Dec 2025"
    )
  ) |>
  ggplot(aes(
    x = occupancy_granted_date,
    y = gross_uplift,
    colour = reflected
  )) +
  geom_point(size = 2, alpha = 0.7) +
  geom_vline(
    xintercept = occupancy_cutoff,
    linetype = "dashed",
    colour = "grey50"
  ) +
  scale_y_continuous(labels = label_dollar(scale = 1e-6, suffix = "M")) +
  scale_x_date(
    date_labels = "%b %Y",
    date_breaks = "6 months",
    guide = guide_axis(angle = 45)
  ) +
  labs(
    title = "Assessment lag: 5 to 8 unit rowhome uplift by occupancy date",
    subtitle = "2024 lots with occupancy granted by Dec 31, 2025; dashed line = Sept 30, 2025",
    x = "Occupancy granted",
    y = "Gross assessed value uplift",
    colour = NULL,
    caption = CAPTION_COE
  ) +
  theme_jd(mode = "light")
```

The measurement cohort for this post is the matched 2024 lots with occupancy granted by September 30, 2025. Everything else, including those October-December completions, goes into the "not yet fully reflected" group.

```{r}
#| label: cohort-split
final_cohort <- rowhome_2024_matched |>
  filter(
    !is.na(occupancy_granted_date),
    occupancy_granted_date <= occupancy_cutoff,
    gross_uplift > 0
  )

not_yet_complete_matched <- rowhome_2024_matched |>
  anti_join(final_cohort, by = "legal_description")

n_final_cohort <- nrow(final_cohort)
```

All together, the measurement cohort is `r n_final_cohort` lots.

#### The rest of the 2024 cohort

The `r n_final_cohort` lots above only cover 2024-permitted lots whose completed buildings the 2026 assessment has caught up with. The remaining matched lots, `r nrow(not_yet_complete_matched)` of them, are either still under construction or finished too late in 2025 for the assessment to reflect. Their 2026 assessed values still tell us something: whatever value the construction had reached, a partial uplift. @tbl-2024-actual-total combines both groups into a single, fully-observed total for the 2024 permit year.

```{r}
#| label: tbl-2024-actual-total
#| tbl-cap: "Actual (not model-based) 2024 cohort assessed value uplift, 5 to 8 unit rowhomes, split by completion status"
status_complete <- "Occupancy by Sept 2025"
status_partial <- "Occupancy after Sept 2025"

actual_2024 <- rowhome_2024_matched |>
  mutate(
    status = if_else(
      legal_description %in% final_cohort$legal_description,
      status_complete,
      status_partial
    )
  ) |>
  select(status, gross_uplift)

actual_2024_summary <- actual_2024 |>
  mutate(status = fct_relevel(status, status_complete, status_partial)) |>
  group_by(status) |>
  summarise(
    n = n(),
    median_gross_uplift = median(gross_uplift),
    total_gross_uplift = sum(gross_uplift),
    .groups = "drop"
  ) |>
  arrange(status) |>
  mutate(status = as.character(status))

gt(actual_2024_summary, rowname_col = "status") |>
  grand_summary_rows(
    fns = "sum",
    fmt = list(
      ~ fmt_currency(
        .,
        columns = c(median_gross_uplift, total_gross_uplift),
        decimals = 1,
        scale_by = 1e-6,
        pattern = "{x}M"
      )
    )
  ) |>
  fmt_currency(
    columns = c(median_gross_uplift, total_gross_uplift),
    decimals = 1,
    scale_by = 1e-6,
    pattern = "{x}M"
  ) |>
  cols_label(
    n = "n",
    median_gross_uplift = "Median gross uplift",
    total_gross_uplift = "Total gross uplift"
  ) |>
  tab_header(
    title = "2024 cohort, actual gross uplift observed so far",
    subtitle = "5 to 8 unit rowhomes; no extrapolation model applied"
  ) |>
  finalize_gt() |>
  opt_stylize_jd()
```

That gives a fully-observed total of `r label_dollars_adaptive(sum(actual_2024_summary$total_gross_uplift))` in assessed value uplift already showing up in the 2026 assessment from the 2024 cohort alone. It understates the eventual value of that cohort: the "not yet fully reflected" group is likely still a stale, mid-construction assessment rather than the finished-building value — exactly the gap the model-based extrapolation later in this post is designed to estimate.

#### From gross uplift to a fairer comparison

The raw 2024-to-2026 difference in assessed value overstates the effect of redevelopment on its own, because Edmonton property values were also generally rising over that period. To isolate the redevelopment effect, I build a comparison: what would each property have been worth in 2026 if it hadn't been redeveloped? I estimate this from a group of RS-zoned properties in mature neighbourhoods that were not part of any 2024-2026 residential new-build permit, taking the median ratio of their 2026 to 2024 assessed value as a general appreciation factor. This un-redeveloped control group may not have appreciated at the rate the redeveloped lots would have, but it's good enough for a blog post.

```{r}
#| label: counterfactual-growth
redeveloped_legal_keys <- bp |>
  st_drop_geometry() |>
  filter(units_added >= 1, work_type == "(01) Building - New") |>
  distinct(legal_description) |>
  pull(legal_description)

nbhd_type_lookup <- bp |>
  st_drop_geometry() |>
  distinct(neighbourhood, neighbourhood_type) |>
  filter(!is.na(neighbourhood_type))

control_group <- assessment_2024 |>
  filter(zoning == "RS") |>
  left_join(nbhd_type_lookup, by = c("neighbourhood_name" = "neighbourhood")) |>
  filter(
    neighbourhood_type == "Mature",
    !legal_description %in% redeveloped_legal_keys,
    !account_number %in% final_cohort$account_number
  ) |>
  inner_join(
    assessment_2026 |> select(account_number, value_2026 = assessed_value),
    by = "account_number"
  ) |>
  mutate(growth_ratio = value_2026 / assessed_value)

g <- median(control_group$growth_ratio, na.rm = TRUE)

# Neighbourhood land-value covariate: median 2024 assessed value of
# RS-zoned properties, per neighbourhood. Used by the single pooled uplift
# model fit across all three small-scale-infill categories below.
nbhd_median_rs_value <- assessment_2024 |>
  filter(zoning == "RS") |>
  group_by(neighbourhood_name) |>
  summarise(nbhd_median_value = median(assessed_value), .groups = "drop")

final_cohort <- final_cohort |>
  mutate(net_uplift = value_2026 - g * value_2024) |>
  left_join(
    nbhd_median_rs_value,
    by = c("neighbourhood" = "neighbourhood_name")
  )

stopifnot(
  all(!is.na(final_cohort$nbhd_median_value)),
  all(!is.na(final_cohort$distance_from_lrt)),
  all(final_cohort$gross_uplift > 0)
)
```

Across `r comma(nrow(control_group))` non-redeveloped RS Zone properties in mature neighbourhoods, the median 2024-to-2026 growth ratio is `r round(g, 2)`x, meaning a typical mature-neighbourhood house that wasn't touched still gained about `r percent(g - 1, accuracy = 1)` in assessed value over those two years. I use this as the baseline against which redevelopment uplift is measured.

This brings us to the two measures of property tax uplift. **Net uplift** is the completed 5 to 8 unit rowhome's 2026 value minus `r round(g, 2)` times its 2024 baseline, isolating the redevelopment effect from general market appreciation. **Gross uplift** is the simple, unadjusted difference, shown alongside it for comparison.

@fig-uplift-distribution shows the distribution of both measures across the `r n_final_cohort` matched lots, split by project size.

```{r}
#| label: fig-uplift-distribution
#| fig-cap: "Distribution of assessed value uplift per redeveloped lot, gross and net of market appreciation"
uplift_long <- final_cohort |>
  select(bucket, gross_uplift, net_uplift) |>
  pivot_longer(
    c(gross_uplift, net_uplift),
    names_to = "type",
    values_to = "uplift"
  ) |>
  mutate(
    type = if_else(
      type == "gross_uplift",
      "Gross uplift",
      "Net of appreciation"
    )
  )

ggplot(uplift_long, aes(x = bucket, y = uplift, fill = type)) +
  geom_boxplot(outlier.alpha = 0.4, position = position_dodge(width = 0.75)) +
  scale_y_continuous(labels = label_dollar(scale = 1e-6, suffix = "M")) +
  labs(
    title = "Assessed value uplift per redeveloped lot",
    subtitle = "2024 RS Zone 5-8 unit rowhomes, mature neighbourhoods,\noccupancy by Sept 2025",
    x = NULL,
    y = "Assessed value uplift",
    fill = NULL,
    caption = CAPTION_COE
  ) +
  theme_jd(mode = "light")
```

@tbl-uplift-rules compares three simple ways of summarizing this uplift into a single per-project figure: a median per lot within two size buckets (5-6 units and 7+ units), a single median per lot across all sizes, and a median per home added, which can be scaled by any project's unit count. These serve as sanity anchors for the model-based rule developed next.

```{r}
#| label: tbl-uplift-rules
#| tbl-cap: "Three ways to summarize per-project uplift for 5 to 8 unit rowhomes, gross and net of appreciation"
rule_bucketed <- final_cohort |>
  group_by(bucket) |>
  summarise(
    n = n(),
    gross = median(gross_uplift),
    net = median(net_uplift),
    .groups = "drop"
  ) |>
  rename(rule = bucket)

rule_pooled <- final_cohort |>
  summarise(n = n(), gross = median(gross_uplift), net = median(net_uplift)) |>
  mutate(rule = "All together", .before = 1)

rule_per_unit <- final_cohort |>
  summarise(
    n = n(),
    gross = median(gross_uplift / units_added),
    net = median(net_uplift / units_added)
  ) |>
  mutate(rule = "Per home added", .before = 1)

bind_rows(rule_bucketed, rule_pooled, rule_per_unit) |>
  gt(rowname_col = "rule") |>
  fmt_currency(
    columns = c(gross, net),
    decimals = 1,
    scale_by = 1e-6,
    pattern = "{x}M"
  ) |>
  cols_label(
    n = "n",
    gross = "Median gross uplift",
    net = "Median net uplift"
  ) |>
  tab_header(
    title = "Three uplift rules compared",
    subtitle = "5 to 8 unit rowhomes, measured from the completed 2024 cohort"
  ) |>
  finalize_gt() |>
  opt_stylize_jd()
```

### Applying the same method to plexes and backyard homes

The same matching-and-consolidation machinery walked through above for the 5 to 8 unit rowhome — legal-key join, 25m geo-fallback, civic-address fallback, same-lot consolidation, split-lot baseline sharing, occupancy cutoff, counterfactual growth adjustment — has already been applied across all three categories together (@sec-consolidating). Backyard homes are structurally different from the other two: they add a suite behind a *retained* house rather than replacing it, so their uplift is a smaller, additive effect rather than a full-lot redevelopment.

```{r}
#| label: other-category-cohorts
# The same measured-cohort recipe used for rowhomes above (occupancy
# cutoff -> net uplift), generalized to any category already consolidated
# into `sites`.
measured_cohort <- function(cat) {
  sites |>
    filter(
      category == cat,
      year == 2024,
      !is.na(value_2024),
      !is.na(value_2026),
      !is.na(occupancy_granted_date),
      occupancy_granted_date <= occupancy_cutoff
    ) |>
    mutate(gross_uplift = value_2026 - value_2024) |>
    filter(gross_uplift > 0) |>
    left_join(
      nbhd_median_rs_value,
      by = c("neighbourhood" = "neighbourhood_name")
    ) |>
    filter(!is.na(nbhd_median_value), !is.na(distance_from_lrt)) |>
    mutate(net_uplift = value_2026 - g * value_2024)
}

plex_final_cohort <- measured_cohort("plex")
backyard_final_cohort <- measured_cohort("backyard")

n_plex_final_cohort <- nrow(plex_final_cohort)
n_backyard_final_cohort <- nrow(backyard_final_cohort)

# No lot's uplift should be counted more than once in either measured
# cohort -- each is already a distinct legal_description by construction,
# so this is a defensive check rather than an expected failure mode.
stopifnot(
  n_distinct(plex_final_cohort$legal_description) == n_plex_final_cohort,
  n_distinct(backyard_final_cohort$legal_description) == n_backyard_final_cohort
)
```

That yields a completed, measured cohort of `r n_plex_final_cohort` 2 to 4 unit plexes and `r n_backyard_final_cohort` backyard homes, alongside the `r n_final_cohort` 5 to 8 unit rowhomes measured above — `r n_final_cohort + n_plex_final_cohort + n_backyard_final_cohort` lots in total to fit a model on.

### A model for uplift {#sec-model-for-uplift}

The simple rules above ignore everything about a project except its unit count. But uplift plausibly varies with location too: land near LRT stations and in more expensive neighbourhoods could command different completed values. So alongside the flat medians, I fit a single model of gross uplift on units added, distance from the nearest LRT stop, the neighbourhood's median RS-zone assessed value, and a backyard-home indicator. I used a generalized additive model (GAM) rather than a plain linear regression for maximum flexibility.

```{r}
#| label: uplift-model
# The pooled cohort every category's model prediction is fit on: the
# measured rowhome cohort plus the plex and backyard cohorts built above,
# with a backyard-home indicator standing in for a full (units_added-
# collinear) project-type factor.
pooled_cohort <- bind_rows(
  final_cohort |> mutate(is_backyard = FALSE),
  plex_final_cohort |> mutate(is_backyard = FALSE),
  backyard_final_cohort |> mutate(is_backyard = TRUE)
)
n_pooled_cohort <- nrow(pooled_cohort)

uplift_model <- gam(
  gross_uplift ~ is_backyard +
    s(units_added, k = 5) +
    s(distance_from_lrt) +
    s(nbhd_median_value),
  data = pooled_cohort,
  method = "REML"
)

# Leave-one-out cross-validated errors: a "good enough" check on the fitted
# GAM against a flat pooled median, not a search over competing specs.
loocv_errs <- function(data) {
  vapply(
    seq_len(nrow(data)),
    function(i) {
      m <- gam(
        gross_uplift ~ is_backyard +
          s(units_added, k = 5) +
          s(distance_from_lrt) +
          s(nbhd_median_value),
        data = data[-i, ],
        method = "REML"
      )
      as.numeric(predict(m, data[i, ])) - data$gross_uplift[i]
    },
    numeric(1)
  )
}
errs_gam <- loocv_errs(pooled_cohort)
rmse_gam <- sqrt(mean(errs_gam^2))
rmse_median <- sqrt(mean(
  (median(pooled_cohort$gross_uplift) - pooled_cohort$gross_uplift)^2
))
bias_by_cat <- tapply(
  errs_gam,
  if_else(
    pooled_cohort$is_backyard,
    "Backyard home",
    if_else(pooled_cohort$units_added <= 4, "2-4plex", "Rowhome")
  ),
  mean
)
```

Checking against @tbl-uplift-rules, the model's average predicted uplift across the measured 5 to 8 unit rowhome cohort alone is `r label_dollars_adaptive(mean(predict(uplift_model, final_cohort |> mutate(is_backyard = FALSE))))` per property, in line with the simple rules. What the model adds is the additional prediction accuracy, which matters when extrapolating to hundreds of lots whose size mix and geography differ from the measured cohort's.

For net uplift, the counterfactual adjustment is applied after prediction: a lot's predicted net uplift is its predicted gross uplift minus (`r round(g, 2) - 1`) times its own 2024 baseline value, the same adjustment used for the measured cohort.

## What all this infill is worth

To go from the measured cohort to the full pipeline, I use the predictions of the model on the non-measured lots: the rest of the 2024 cohort (lots that didn't match or aren't fully assessed yet), and lots from 2025 and 2026.

::: {.callout-note}
Model predictions are **steady-state estimates** of uplift. They represent the eventual outcome once construction wraps up and a future assessment incorporates the finished product.
:::


```{r}
#| label: pipeline-counts
pipeline_counts <- sites |>
  filter(category == "rowhome") |>
  mutate(bucket = if_else(units_added <= 6, "5-6 units", "7+ units")) |>
  count(year, bucket)

# H1 annualization multiplier, same method as the Q2 2026 report: median
# ratio of full-year to H1 residential units across historical years.
h1_ratios <- bp |>
  st_drop_geometry() |>
  filter(year < 2026) |>
  group_by(year) |>
  summarize(
    full_year = sum(units_added),
    h1 = sum(units_added[month_number <= 6]),
    .groups = "drop"
  ) |>
  mutate(ratio = full_year / h1)
h1_multiplier <- round(median(h1_ratios$ratio), 1)

n_2024_all <- pipeline_counts |>
  filter(year == 2024) |>
  summarise(n = sum(n)) |>
  pull(n)
n_2025_all <- pipeline_counts |>
  filter(year == 2025) |>
  summarise(n = sum(n)) |>
  pull(n)
n_2026_h1 <- pipeline_counts |>
  filter(year == 2026) |>
  summarise(n = sum(n)) |>
  pull(n)
n_2026_projected <- n_2026_h1 * h1_multiplier
```

```{r}
#| label: extrapolation
# Every site's model prediction, whatever its category or year -- a
# single table the rest of this post's extrapolation reads from. A site
# is "measured" if it's in one of the three measured cohorts already built
# above (checked by legal_description, each site's unique key); everything
# else gets a prediction from the pooled model, using its own baseline
# where matched or its neighbourhood's median RS value as a stand-in
# where not.
measured_legal_keys <- c(
  final_cohort$legal_description,
  plex_final_cohort$legal_description,
  backyard_final_cohort$legal_description
)

median_distance <- median(sites$distance_from_lrt, na.rm = TRUE)
median_nbhd_value <- median(nbhd_median_rs_value$nbhd_median_value)

all_sites_pred <- sites |>
  filter(year %in% 2024:2026) |>
  left_join(
    nbhd_median_rs_value,
    by = c("neighbourhood" = "neighbourhood_name")
  ) |>
  mutate(
    nbhd_median_value = coalesce(nbhd_median_value, median_nbhd_value),
    distance_from_lrt = coalesce(distance_from_lrt, median_distance),
    baseline_value = coalesce(value_2024, nbhd_median_value),
    is_backyard = category == "backyard",
    is_measured = legal_description %in% measured_legal_keys,
    pred_gross = as.numeric(predict(
      uplift_model,
      newdata = pick(
        is_backyard,
        units_added,
        distance_from_lrt,
        nbhd_median_value
      )
    )),
    pred_net = pred_gross - (g - 1) * baseline_value
  )

# Blended municipal tax rate per rowhome size bucket, weighted by each
# bucket's observed mix of Residential vs. Other Residential assessment
# class in the measured cohort. Used both for the extrapolated tax figures
# below and, via bucket_rates_2026, for the neighbourhood breakdown later.
bucket_rate <- function(bucket_data, yr) {
  bucket_data |>
    mutate(bucket = if_else(units_added <= 6, "5-6 units", "7+ units")) |>
    count(bucket, mill_class_1) |>
    group_by(bucket) |>
    mutate(share = n / sum(n)) |>
    summarise(
      rate = sum(
        share *
          if_else(
            mill_class_1 == "OTHER RESIDENTIAL",
            municipal_rate("Other Residential", yr),
            municipal_rate("Residential", yr)
          )
      ),
      .groups = "drop"
    )
}

bucket_rates_2026 <- bucket_rate(final_cohort, 2026)
bucket_rates_2025 <- bucket_rate(final_cohort, 2025)

# A category's own flat blended rate (plex/backyard cohorts are too small
# to split further by size).
category_tax_rate <- function(cohort, yr) {
  cohort |>
    count(mill_class_1) |>
    mutate(share = n / sum(n)) |>
    summarise(
      rate = sum(
        share *
          if_else(
            mill_class_1 == "OTHER RESIDENTIAL",
            municipal_rate("Other Residential", yr),
            municipal_rate("Residential", yr)
          )
      )
    ) |>
    pull(rate)
}
plex_rate_2026 <- category_tax_rate(plex_final_cohort, 2026)
backyard_rate_2026 <- category_tax_rate(backyard_final_cohort, 2026)

all_sites_pred <- all_sites_pred |>
  mutate(
    bucket = if_else(units_added <= 6, "5-6 units", "7+ units"),
    rate_2026 = case_when(
      category == "rowhome" ~ NA_real_,
      category == "plex" ~ plex_rate_2026,
      category == "backyard" ~ backyard_rate_2026
    ),
    rate_2025 = if_else(category == "plex", plex_rate_2026, backyard_rate_2026)
  ) |>
  left_join(bucket_rates_2026, by = "bucket") |>
  mutate(rate_2026 = coalesce(rate_2026, rate)) |>
  select(-rate) |>
  left_join(bucket_rates_2025, by = "bucket") |>
  mutate(rate_2025 = if_else(category == "rowhome", rate, rate_2025)) |>
  select(-rate) |>
  mutate(
    pred_tax_gross_2026 = pred_gross * rate_2026,
    pred_tax_net_2026 = pred_net * rate_2026,
    pred_tax_net_2025 = pred_net * rate_2025
  )

# Summarizes one category's steady-state total: measured sites keep their
# actual net uplift; everything else uses the model prediction, weighted
# by the H1-2026 annualization multiplier for the un-permitted remainder
# of 2026.
category_totals <- function(cat, measured) {
  pred <- all_sites_pred |> filter(category == cat)

  model_2024 <- pred |> filter(year == 2024, !is_measured)
  model_2025 <- pred |> filter(year == 2025)
  model_2026_h1 <- pred |> filter(year == 2026)

  sum_cols <- function(d) {
    d |>
      summarise(
        gross = sum(pred_gross),
        net = sum(pred_net),
        tax_gross_2026 = sum(pred_tax_gross_2026),
        tax_net_2026 = sum(pred_tax_net_2026),
        tax_net_2025 = sum(pred_tax_net_2025)
      )
  }
  m2024 <- sum_cols(model_2024)
  m2025 <- sum_cols(model_2025)
  m2026h1 <- sum_cols(model_2026_h1)
  m2026rest <- m2026h1 |>
    mutate(across(everything(), \(x) x * (h1_multiplier - 1)))

  measured_totals <- measured |>
    summarise(gross = sum(gross_uplift), net = sum(net_uplift))

  n_2024 <- pred |> filter(year == 2024) |> nrow()
  n_2025 <- pred |> filter(year == 2025) |> nrow()
  n_2026_h1_n <- pred |> filter(year == 2026) |> nrow()
  homes_2024 <- pred |>
    filter(year == 2024) |>
    summarise(u = sum(units_added)) |>
    pull(u)
  homes_2025 <- pred |>
    filter(year == 2025) |>
    summarise(u = sum(units_added)) |>
    pull(u)
  homes_2026_h1 <- pred |>
    filter(year == 2026) |>
    summarise(u = sum(units_added)) |>
    pull(u)

  list(
    n_projects = n_2024 + n_2025 + round(n_2026_h1_n * h1_multiplier),
    homes_added = homes_2024 +
      homes_2025 +
      round(homes_2026_h1 * h1_multiplier),
    total_gross_uplift = measured_totals$gross +
      m2024$gross +
      m2025$gross +
      m2026h1$gross +
      m2026rest$gross,
    total_net_uplift = measured_totals$net +
      m2024$net +
      m2025$net +
      m2026h1$net +
      m2026rest$net,
    annual_tax_uplift_gross = sum(
      measured |>
        mutate(
          new_rate = if_else(
            mill_class_1 == "OTHER RESIDENTIAL",
            municipal_rate("Other Residential", 2026),
            municipal_rate("Residential", 2026)
          ),
          t = new_rate *
            value_2026 -
            municipal_rate("Residential", 2026) * value_2024
        ) |>
        pull(t)
    ) +
      m2024$tax_gross_2026 +
      m2025$tax_gross_2026 +
      m2026h1$tax_gross_2026 +
      m2026rest$tax_gross_2026,
    annual_tax_uplift_net = sum(
      measured |>
        mutate(
          new_rate = if_else(
            mill_class_1 == "OTHER RESIDENTIAL",
            municipal_rate("Other Residential", 2026),
            municipal_rate("Residential", 2026)
          ),
          t = new_rate *
            value_2026 -
            municipal_rate("Residential", 2026) * g * value_2024
        ) |>
        pull(t)
    ) +
      m2024$tax_net_2026 +
      m2025$tax_net_2026 +
      m2026h1$tax_net_2026 +
      m2026rest$tax_net_2026,
    model_based_2024 = m2024,
    model_based_2025 = m2025,
    model_based_2026_h1 = m2026h1,
    model_based_2026_rest = m2026rest,
    measured_totals = measured_totals
  )
}

rowhome_results <- category_totals("rowhome", final_cohort)
plex_results <- category_totals("plex", plex_final_cohort)
backyard_results <- category_totals("backyard", backyard_final_cohort)
```

```{r}
#| label: comprehensive-extrapolation
comprehensive_summary <- bind_rows(
  tibble(
    category = "5-8 unit rowhome",
    n_projects = rowhome_results$n_projects,
    homes_added = rowhome_results$homes_added,
    net_uplift = rowhome_results$total_net_uplift,
    annual_tax_uplift = rowhome_results$annual_tax_uplift_net
  ),
  tibble(
    category = "2-4 unit plex",
    n_projects = plex_results$n_projects,
    homes_added = plex_results$homes_added,
    net_uplift = plex_results$total_net_uplift,
    annual_tax_uplift = plex_results$annual_tax_uplift_net
  ),
  tibble(
    category = "Backyard home",
    n_projects = backyard_results$n_projects,
    homes_added = backyard_results$homes_added,
    net_uplift = backyard_results$total_net_uplift,
    annual_tax_uplift = backyard_results$annual_tax_uplift_net
  )
)
comprehensive_summary <- bind_rows(
  comprehensive_summary,
  comprehensive_summary |>
    summarise(
      n_projects = sum(n_projects),
      homes_added = sum(homes_added),
      net_uplift = sum(net_uplift),
      annual_tax_uplift = sum(annual_tax_uplift)
    ) |>
    mutate(category = "Total", .before = 1)
)

comprehensive_net_uplift <- comprehensive_summary |>
  filter(category == "Total") |>
  pull(net_uplift)
comprehensive_tax_uplift <- comprehensive_summary |>
  filter(category == "Total") |>
  pull(annual_tax_uplift)
rowhome_totals <- list(net_uplift = rowhome_results$total_net_uplift)
```

@fig-comprehensive-infill plots the net value uplift side by side. The 5 to 8 unit rowhome dominates, both because it adds the most units per lot and because it's been the more common redevelopment path since ZBR, but 2 to 4 unit plexes add a meaningful share on their own. Backyard homes do not add a significant amount in assessment base uplift due to their smaller scale.

```{r}
#| label: fig-comprehensive-infill
#| fig-cap: "Net assessed value uplift by small-scale infill category, 2024 through H1 2026 (2026 full year projected)"
comprehensive_summary |>
  filter(category != "Total") |>
  mutate(category = fct_reorder(category, net_uplift)) |>
  ggplot(aes(x = category, y = net_uplift, fill = category)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_y_continuous(labels = label_dollar(scale = 1e-6, suffix = "M")) +
  labs(
    title = "Net value uplift by small-scale infill category",
    subtitle = "Net of market appreciation; 2026 full year projected",
    x = NULL,
    y = "Net assessed value uplift",
    caption = CAPTION_COE
  ) +
  theme_jd(mode = "light")
```

Estimating the corresponding tax revenue in @tbl-comprehensive-infill, we can see that RS Zone infill will bring in an estimated **`r label_dollars_adaptive(comprehensive_tax_uplift)`** in additional municipal tax revenue at a steady state.

```{r}
#| label: tbl-comprehensive-infill
#| tbl-cap: "RS Zone infill uplift, 2024 through H1 2026 (2026 full year projected)"
comprehensive_summary |>
  filter(category != "Total") |>
  gt(rowname_col = "category") |>
  fmt_currency(
    columns = c(net_uplift, annual_tax_uplift),
    decimals = 1,
    scale_by = 1e-6,
    pattern = "{x}M"
  ) |>
  fmt_integer(columns = c(n_projects, homes_added)) |>
  grand_summary_rows(
    fns = "sum",
    fmt = list(
      ~ fmt_integer(., columns = c(n_projects, homes_added)),
      ~ fmt_currency(
        .,
        columns = c(net_uplift, annual_tax_uplift),
        decimals = 1,
        scale_by = 1e-6,
        pattern = "{x}M"
      )
    )
  ) |>
  cols_label(
    n_projects = "Lots",
    homes_added = "Homes added",
    net_uplift = "Net value uplift",
    annual_tax_uplift = "Annual municipal tax uplift"
  ) |>
  tab_header(
    title = "RS Zone infill property assessment and municipal tax uplift",
    subtitle = "2024 through H1 2026 (2026 full year projected)"
  ) |>
  finalize_gt() |>
  opt_stylize_jd()
```

That's not insignificant. The infill tax uplift would cover a bunch of the items in the [2026 operating budget](https://www.edmonton.ca/sites/default/files/public-files/documents/2026-approved-operating-budget-april.pdf) that materially improve Edmontonians' quality of life:

- **DATS Service Level ($3.0M)** — DATS (Dedicated Accessible Transit Service) is the door-to-door transit option for Edmontonians who can't use conventional buses or LRT. Demand had outgrown the existing budget; without this funding, the share of trip requests the city could confirm was projected to fall from 98% to 89%, denying an estimated 120,000 rider trips a year.
- **Community Standards Peace Officers Contingent ($5.8M)** — funds a new 31-person Traffic Safety Team (phased in through 2027) focused on school-zone, speeding, and vehicle-noise enforcement.
- **Permanent Enhanced Cleaning ($2.7M)** — makes permanent a temporary boost to cleaning of transit stations and vehicles, aimed at rider safety and comfort.
- **Bus Cleaning Service Delivery Model ($1.2M)** — restores a bus-cleaning budget cut made in 2020, when the City briefly outsourced the work; the outsourcing decision was reversed the next year, but the cut to the cleaning budget itself was never reinstated until now.
- **Infill Liaison Team ($0.6M)** — reinstates a dedicated team to field infill-related complaints, explain the City's infill policy, and coordinate between departments. A fitting one to fund with infill's own tax revenue.
- **Facility Maintenance Services – Contract Work Inflation ($0.6M)** — covers inflation-driven cost increases (about 15%) on contracted maintenance work at City Hall, recreation centres, police stations, and libraries.

Alternatively, this infill tax uplift is equivalent to an approximate 0.6% reduction in the property tax rate, which would be a welcome relief for all property owners in Edmonton. As we build more small-scale infill in 2027 and beyond, the tax uplift will continue to grow, further decreasing property taxes for everyone. [Based on $11M for Explore Edmonton [worth 0.5% of property tax rate](https://edmontonjournal.com/news/local-news/explore-edmonton-celebrates-funding-as-city-vies-for-mystery-event)]{.aside}

```{r}
#| label: key-facts-validation
#| include: false
# Sanity check that the hardcoded Key facts bullets above still match the
# computed figures, so a future re-render can't silently drift from them.
stopifnot(
  comprehensive_net_uplift > 1.5e9,
  comprehensive_net_uplift < 1.9e9,
  comprehensive_tax_uplift > 1.2e7,
  comprehensive_tax_uplift < 1.5e7
)
```

## Conclusions

The RS Zone's small-scale infill isn't just adding homes, it's increasing the size of the property tax base. A few hundred developments over the past two and a half years, built in mature areas that already have the pipes, roads, and schools in place, is already worth more than `r label_dollars_adaptive(comprehensive_tax_uplift)` a year in additional municipal property tax capacity at steady state. This corresponds to a 0.6% decrease in the property tax rate, which is a meaningful relief for all property owners in Edmonton. And more relief will be on the way as more gets built in 2027 and onwards.

Now it's up to Council to decide what to do with the ever-growing infill dividend.
Whether it's used to keep property taxes lower or to increase service levels, the Zoning Bylaw Renewal will continue to pay off for all Edmontonians for years to come.
If you'd like to help keep it that way, join me at [Grow Together Edmonton](https://www.growtogetheryeg.com/) where we advocate for building more housing options in the places people want to live.
