---
title: "The property tax dividend from Edmonton's RS Zone rowhomes"
date: "2026-07-12"
author: "Jacob Dawang"
categories: [housing, edmonton, zoning]
description: "Redeveloping an old house into a 5-8 unit RS Zone rowhome doesn't just add homes, it multiplies the property's assessed value. Here's what that's worth to Edmonton's property tax base since the Zoning Bylaw Renewal took effect."
link-external-newwindow: true
link-external-icon: false
execute:
  cache: false
draft: false
---

Edmonton's Zoning Bylaw Renewal (ZBR) has spent two and a half years adding new homes to existing neighbourhoods via the extremely successful 5-to-8-unit rowhome. [In an ongoing series of blog posts](../q2-building-permits/index.qmd), I track how the RS zone is slowly transforming our city. This post asks a different question: what is that redevelopment worth to the City's property tax base?

When a single-detached home in a mature neighbourhood becomes a RS Zone rowhome, its assessed value increases, adding to the City's property tax revenue.

Using building permits from 2024, the first year the RS Zone was in force, matched against the City's property assessment data, I measure that uplift directly for completed projects, then use a regression model fit on those measurements to estimate what it adds up to across the full 2024-2026 pipeline.

## Key facts

-   Mature-neighbourhood RS Zone rowhomes (5 to 8 units) will add an estimated **$1.3 billion** in assessed property value at steady state — once every 2024-2026 project is complete and fully reflected in an assessment roll — net of general market appreciation, across roughly **800** permitted redevelopment projects.

-   At 2026 municipal tax rates, that steady-state uplift will generate on the order of **$10-11 million** in property tax every year, a permanent addition to the tax base rather than a one-time bump.

-   Because most of that pipeline is still under construction or was only recently completed, the City has collected only about **$2 million** in additional property tax so far, across 2025 and 2026 combined — far below the eventual run rate.

-   No subsidy, grant, or City-led development produced this. Removing rules that made this kind of building illegal did.

## Data

This analysis combines three of Edmonton's Open Data sources. Edmonton's [general building permit data](https://data.edmonton.ca/Urban-Planning-Economy/General-Building-Permits/24uj-dj8v/about_data), the same source used in the [Q2 2026 building permit report](../q2-building-permits/index.qmd), identifies which properties were redeveloped into 5-8 unit RS Zone rowhomes and when. Edmonton's [current calendar year property assessment data](https://data.edmonton.ca/City-Administration/Property-Assessment-Data-Current-Calendar-Year-/q7d6-ambg/about_data) gives each property's assessed value as of 2026, and Edmonton's [historical property assessment data](https://data.edmonton.ca/City-Administration/Property-Assessment-Data-Historical-/qi6a-xuwt/about_data) gives assessed values back to 2012, including 2024 and 2025.

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(sf)
library(gt)
library(scales)
library(mountainmathHelpers)
library(jdawangHelpers)

options(
  cancensus.api_key = Sys.getenv("CM_API_KEY"),
  cancensus.cache_path = Sys.getenv("CM_CACHE_PATH"),
  tongfen.cache_path = Sys.getenv("TONGFEN_CACHE_PATH"),
  nextzen_API_key = Sys.getenv("NEXTZEN_API_KEY")
)

label_dollars_adaptive <- function(x) {
  if (abs(x) >= 1e9) {
    label_dollar(scale = 1e-9, suffix = "B", accuracy = 0.1)(x)
  } else if (abs(x) >= 1e8) {
    label_dollar(scale = 1e-6, suffix = "M", accuracy = 1)(x)
  } else {
    label_dollar(scale = 1e-6, suffix = "M", accuracy = 0.1)(x)
  }
}

bp_raw <- get_edmonton_building_permit_data(
  cache_dir = Sys.getenv("EDMONTON_BP_CACHE_PATH"),
  end_date = "2026-06-30"
)
crs <- lambert_conformal_conic_at(bp_raw)

bp <- bp_raw |>
  st_transform(crs) |>
  mutate(permit_row_id = row_number()) |>
  filter(year < 2026 | (year == 2026 & month_number <= 6)) |>
  filter_edmonton_residential() |>
  filter(units_added >= 1, work_type != "(04) Excavation") |>
  add_edmonton_project_type() |>
  add_edmonton_suite_info() |>
  add_edmonton_neighbourhood_type()

# LRT stops (existing + under construction), same setup as the Q2 2026 report
future_stops <- read_sf("../../../data/future_lrt_stops.geojson") |>
  filter(
    !(stop_name %in%
      c(
        "Castle Downs Station",
        "145 Ave Station",
        "137 Ave Station",
        "132 Ave Station"
      ))
  )

transit_stops <- load_edmonton_transit_stops(
  gtfs_path = "../../../data/ca-alberta-edmonton-transit-system-gtfs-714.zip",
  service_date = as.Date("2023-11-09"),
  future_stops = future_stops,
  crs = crs
)

# The pipeline this whole post is about: mature-neighbourhood RS Zone
# 5-8 unit new-builds, permitted since ZBR took effect. Each permit gets
# its distance to the nearest LRT stop, used by the uplift model later.
rowhome_pipeline <- bp |>
  filter(
    zoning == "RS",
    between(units_added, 5, 8),
    work_type == "(01) Building - New",
    neighbourhood_type == "Mature"
  ) |>
  add_transit_distance(transit_stops)
```

```{r}
#| label: fetch-assessment-data
# This is a one-off fetch specific to this post: Edmonton's property
# assessment data hasn't been used on this blog before, so this helper
# isn't (yet) part of jdawangHelpers.
fetch_socrata_csv <- function(resource, query, cache_file) {
  cache_dir <- Sys.getenv("EDMONTON_BP_CACHE_PATH")
  path <- file.path(cache_dir, cache_file)
  if (!file.exists(path)) {
    url <- paste0(
      "https://data.edmonton.ca/resource/",
      resource,
      ".csv?",
      URLencode(query)
    )
    download.file(url, path, mode = "wb")
  }
  read_csv(path, show_col_types = FALSE, col_types = cols(.default = "c"))
}

assessment_hist <- fetch_socrata_csv(
  "qi6a-xuwt",
  "$where=assessment_year in (2023,2024,2025)&$limit=1000000",
  "assessment_2023_2025.csv"
) |>
  mutate(
    assessment_year = as.integer(assessment_year),
    assessed_value = as.numeric(assessed_value),
    latitude = as.numeric(latitude),
    longitude = as.numeric(longitude)
  )

assessment_2026 <- fetch_socrata_csv(
  "q7d6-ambg",
  "$limit=600000",
  "assessment_2026.csv"
) |>
  mutate(
    assessed_value = as.numeric(assessed_value),
    latitude = as.numeric(latitude),
    longitude = as.numeric(longitude)
  )

tax_rates <- fetch_socrata_csv("pwis-wc4c", "$limit=10000", "tax_rates.csv") |>
  mutate(
    tax_year = as.integer(tax_year),
    amount_per_dollar_of_assessed_value = as.numeric(
      amount_per_dollar_of_assessed_value
    )
  )

assessment_2023 <- assessment_hist |> filter(assessment_year == 2023)
assessment_2024 <- assessment_hist |> filter(assessment_year == 2024)
assessment_2025 <- assessment_hist |> filter(assessment_year == 2025)

municipal_rate <- function(class, yr) {
  tax_rates |>
    filter(
      tax_year == yr,
      tax_rate_type == "Municipal",
      assessment_class == class
    ) |>
    pull(amount_per_dollar_of_assessed_value)
}
```

## Measuring the uplift: the 2024 cohort

The 2026 property assessment is supposed to reflect each property's physical condition as of December 31, 2025. In principle, that means any 2024 permit whose building had received occupancy by the end of 2025 should be assessed as completed in the 2026 assessment roll. In practice, the assessments lag, as @sec-assessment-catch-up checks directly.

```{r}
#| label: measurement-cohort
cohort_2024 <- rowhome_pipeline |>
  filter(year == 2024)

n_cohort_2024_all <- nrow(cohort_2024)
n_occupancy_by_dec <- cohort_2024 |>
  st_drop_geometry() |>
  filter(
    !is.na(occupancy_granted_date),
    occupancy_granted_date <= as.Date("2025-12-31")
  ) |>
  nrow()
```

Edmonton issued **`r n_cohort_2024_all`** mature-neighbourhood RS Zone 5-8 unit new-build permits in 2024, of which **`r n_occupancy_by_dec`** had received occupancy by the end of 2025. I match all `r n_cohort_2024_all` to their assessment records, then use the occupancy dates to decide which ones the 2026 assessment actually reflects as complete.

### Matching permits to assessments

The key challenge is linking a specific building permit to the same physical property, before and after redevelopment. Account numbers usually persist through redevelopment, but addresses often don't: a lot can get re-addressed when it's rebuilt, so a permit's civic address is frequently not the address that appears in the property's earlier assessment record. Instead, I match on legal description (plan/block/lot), which is stable across redevelopment, with a fallback to geographic proximity for the small number of properties where the legal description format doesn't parse identically on both sides, or where redevelopment came with a brand new account number.

I parse each side's legal description into a `(plan, block, lot)` key and join on it. A building permit's legal description looks like `Plan 715HW Blk 25 Lot 22`; the historical assessment data records the same parcel as `Plan: 715HW  Block: 25  Lot: 22`.

```{r}
#| label: legal-key-matching
parse_permit_legal <- function(x) {
  m <- str_match(x, "Plan\\s+(\\S+)\\s+Blk\\s+(\\S+)\\s+Lots?\\s+(\\S+)")
  tibble(plan = m[, 2], block = m[, 3], lot = m[, 4])
}
parse_assessment_legal <- function(x) {
  m <- str_match(x, "Plan:\\s*(\\S+)\\s+Block:\\s*(\\S+)\\s+Lot:\\s*(\\S+)")
  tibble(plan = m[, 2], block = m[, 3], lot = m[, 4])
}
normalize_key <- function(plan, block, lot) {
  paste(
    toupper(str_trim(plan)),
    toupper(str_trim(block)),
    toupper(str_trim(lot)),
    sep = "|"
  )
}

assessment_2024_keyed <- assessment_2024 |>
  bind_cols(parse_assessment_legal(assessment_2024$legal_description)) |>
  mutate(key = normalize_key(plan, block, lot)) |>
  filter(!is.na(key)) |>
  distinct(key, .keep_all = TRUE)

# Reusable matcher: join a cohort of permits to their 2024 baseline assessed
# value by legal-description key. Used for the 2024 cohort here, and again
# later for the 2025/2026 permits when estimating net uplift.
match_baseline_2024 <- function(cohort_sf) {
  cohort_sf |>
    st_drop_geometry() |>
    bind_cols(parse_permit_legal(cohort_sf$legal_description)) |>
    mutate(key = normalize_key(plan, block, lot)) |>
    left_join(
      assessment_2024_keyed |>
        select(key, account_number, value_2024 = assessed_value),
      by = "key"
    )
}

cohort_matched <- match_baseline_2024(cohort_2024)
n_matched_legal <- sum(!is.na(cohort_matched$account_number))
```

That legal-key match links most properties to their 2024 baseline assessed value directly. I close the gap two ways: a 25m geographic fallback for lots consolidated from two legal titles into one, and hand-matching for the handful of re-platted lots that geography can't reach either.

```{r}
#| label: geo-fallback-2024
# Reusable: fall back to the nearest 2024 assessment point within 25m for
# any permit the legal-key match missed.
geo_fallback_baseline <- function(matched_df, cohort_sf) {
  still_unmatched <- matched_df |> filter(is.na(account_number))
  if (nrow(still_unmatched) == 0) {
    return(matched_df)
  }

  unmatched_sf <- cohort_sf |>
    filter(permit_row_id %in% still_unmatched$permit_row_id)
  assessment_2024_sf <- assessment_2024 |>
    filter(!is.na(latitude), !is.na(longitude)) |>
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
    st_transform(crs)

  nearest_idx <- st_nearest_feature(unmatched_sf, assessment_2024_sf)
  dist_m <- as.numeric(st_distance(
    unmatched_sf,
    assessment_2024_sf[nearest_idx, ],
    by_element = TRUE
  ))

  geo_matches <- tibble(
    permit_row_id = unmatched_sf$permit_row_id,
    geo_account = assessment_2024_sf$account_number[nearest_idx],
    geo_value_2024 = assessment_2024_sf$assessed_value[nearest_idx],
    geo_dist_m = dist_m
  ) |>
    filter(geo_dist_m <= 25)

  matched_df |>
    left_join(geo_matches, by = "permit_row_id") |>
    mutate(
      account_number = coalesce(account_number, geo_account),
      value_2024 = coalesce(value_2024, geo_value_2024)
    ) |>
    select(-any_of(c("geo_account", "geo_value_2024", "geo_dist_m")))
}

cohort_matched <- geo_fallback_baseline(cohort_matched, cohort_2024)

n_matched_after_geo <- sum(!is.na(cohort_matched$account_number))
n_still_dropped <- n_cohort_2024_all - n_matched_after_geo
```

The 25m fallback closes most of the remaining gap — in each case to the exact same civic address as the permit, at a matched distance of a few millimetres. The last `r n_still_dropped` properties carry a legal description reissued under a brand-new subdivision plan and have no permit coordinates, so neither match reaches them. I hand-matched these by tracing each address through the historical assessment data; two had no 2024 assessment under either their old or new legal description, so I escalate their last observed 2023 value to a 2024 basis using the same median growth-ratio approach used elsewhere in this post. [I am not sure why I cannot find these two properties in the 2024 assessment roll. If you have any ideas, let me know. The addresses are 13009 - 109A AVENUE NW and 1, 11630 - 122 STREET NW]{.aside}

```{r}
#| label: manual-matches
# Three 2024 new-build permits carry re-platted (post-subdivision) legal
# descriptions and no permit coordinates, so neither the legal-key match nor
# the 25m geo-fallback finds them. Hand-matched by tracing each address's
# assessment history: two properties were demolished before the 2024 roll's
# snapshot date and only have a 2023 assessed value, escalated to a 2024
# basis by the median 2023->2024 growth ratio among parcels present in both
# rolls (same idiom as the 2024->2026 counterfactual growth ratio used later
# in this post). Zoning isn't used to restrict this comparison: 2023 predates
# ZBR, so its zoning codes are the pre-ZBR system (e.g. RF1, RF3) rather than
# "RS", and the account_number join already restricts to persisting parcels.
g23 <- assessment_2023 |>
  inner_join(
    assessment_2024 |> select(account_number, v24 = assessed_value),
    by = "account_number"
  ) |>
  mutate(ratio = v24 / assessed_value) |>
  filter(assessed_value > 10000, ratio > 0.5, ratio < 3) |>
  summarise(g = median(ratio, na.rm = TRUE)) |>
  pull(g)

lookup_hist_value <- function(acct, yr) {
  assessment_hist |>
    filter(account_number == acct, assessment_year == yr) |>
    pull(assessed_value)
}

manual_matches <- tribble(
  ~legal_description           , ~baseline_account , ~baseline_year , ~outcome_account ,
  "Plan 2420304 Blk 1 Lot 10C" , "2696003"         , 2023L          , "11222239"       ,
  "Plan 2421625 Blk 8 Lot 4A"  , "2122703"         , 2024L          , "11311520"       ,
  "Plan 2421483 Blk 35 Lot 8C" , "2775609"         , 2023L          , "11229512"
) |>
  rowwise() |>
  mutate(baseline_raw = lookup_hist_value(baseline_account, baseline_year)) |>
  ungroup() |>
  mutate(
    value_2024_manual = if_else(
      baseline_year == 2023L,
      baseline_raw * g23,
      baseline_raw
    )
  )

cohort_matched <- cohort_matched |>
  left_join(
    manual_matches |>
      select(
        legal_description,
        manual_account = outcome_account,
        manual_value_2024 = value_2024_manual
      ),
    by = "legal_description"
  ) |>
  mutate(
    account_number = coalesce(account_number, manual_account),
    value_2024 = coalesce(value_2024, manual_value_2024)
  ) |>
  select(-manual_account, -manual_value_2024)

n_matched_total <- sum(!is.na(cohort_matched$account_number))
n_dropped <- n_cohort_2024_all - n_matched_total
```

Combining the geo-fallback with these three hand-matched properties links all **`r n_matched_total`** of `r n_cohort_2024_all` permits (`r percent(n_matched_total / n_cohort_2024_all, accuracy = 1)`) to a 2024 baseline assessed value. In most cases the account number persists through redevelopment, so I use it to look up the completed building's 2026 assessed value, again with a geographic fallback for the small number of properties issued a brand new account number when redeveloped.

```{r}
#| label: join-2026-and-2025
# Reusable: join a baseline-matched cohort to its 2026 outcome value by
# account_number, falling back to the nearest 2026 assessment point within
# 25m for the rare case where redevelopment came with a new account number.
match_outcome_2026 <- function(matched_df, cohort_sf) {
  joined <- matched_df |>
    filter(!is.na(account_number)) |>
    left_join(
      assessment_2026 |>
        select(account_number, value_2026 = assessed_value, mill_class_1),
      by = "account_number"
    )

  missing_2026 <- joined |> filter(is.na(value_2026))
  if (nrow(missing_2026) == 0) {
    return(joined)
  }

  missing_2026_sf <- cohort_sf |>
    filter(permit_row_id %in% missing_2026$permit_row_id)
  assessment_2026_sf <- assessment_2026 |>
    filter(!is.na(latitude), !is.na(longitude)) |>
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
    st_transform(crs)

  nearest_idx <- st_nearest_feature(missing_2026_sf, assessment_2026_sf)
  geo_2026 <- tibble(
    permit_row_id = missing_2026_sf$permit_row_id,
    geo_value_2026 = assessment_2026_sf$assessed_value[nearest_idx],
    geo_mill_class = assessment_2026_sf$mill_class_1[nearest_idx],
    geo_dist_m = as.numeric(st_distance(
      missing_2026_sf,
      assessment_2026_sf[nearest_idx, ],
      by_element = TRUE
    ))
  ) |>
    filter(geo_dist_m <= 25)

  joined |>
    left_join(geo_2026, by = "permit_row_id") |>
    mutate(
      value_2026 = coalesce(value_2026, geo_value_2026),
      mill_class_1 = coalesce(mill_class_1, geo_mill_class)
    ) |>
    select(-any_of(c("geo_value_2026", "geo_mill_class", "geo_dist_m")))
}

# 2025 (mid-construction) value, used later for the partial-uplift estimate
matched_2024_all <- match_outcome_2026(cohort_matched, cohort_2024) |>
  left_join(
    assessment_2025 |> select(account_number, value_2025 = assessed_value),
    by = "account_number"
  ) |>
  filter(!is.na(value_2026)) |>
  mutate(
    bucket = if_else(units_added <= 6, "5-6 units", "7-8 units"),
    gross_uplift = value_2026 - value_2024
  )

n_matched_2024_all <- nrow(matched_2024_all)
```

That yields **`r n_matched_2024_all`** 2024 permits with both a 2024 baseline and a 2026 assessed value.

### When does the assessment catch up? {#sec-assessment-catch-up}

Before treating every occupancy-granted building as "complete" in the 2026 data, it's worth checking whether the assessments actually keep up with late-year completions. @fig-assessment-lag plots each matched property's gross uplift against its occupancy date. Buildings that received occupancy by roughly the end of September 2025 show the full rowhome uplift. For many October, November, and especially December completions, the uplift is much smaller and doesn't follow the same pattern — probably because their assessments still reflect an unfinished state.

```{r}
#| label: fig-assessment-lag
#| fig-cap: "Gross assessed value uplift by occupancy date: the 2026 assessment only reliably reflects buildings finished by about September 2025"
occupancy_cutoff <- as.Date("2025-09-30")

matched_2024_all |>
  filter(
    !is.na(occupancy_granted_date),
    occupancy_granted_date <= as.Date("2025-12-31")
  ) |>
  mutate(
    reflected = if_else(
      occupancy_granted_date <= occupancy_cutoff,
      "Occupancy by Sept 2025",
      "Occupancy Oct-Dec 2025"
    )
  ) |>
  ggplot(aes(
    x = occupancy_granted_date,
    y = gross_uplift,
    colour = reflected
  )) +
  geom_point(size = 2, alpha = 0.7) +
  geom_vline(
    xintercept = occupancy_cutoff,
    linetype = "dashed",
    colour = "grey50"
  ) +
  scale_y_continuous(labels = label_dollar(scale = 1e-6, suffix = "M")) +
  scale_x_date(
    date_labels = "%b %Y",
    date_breaks = "6 months",
    guide = guide_axis(angle = 45)
  ) +
  labs(
    title = "Assessment lag: uplift by occupancy date",
    subtitle = "2024 permits with occupancy granted by Dec 31, 2025; dashed line = Sept 30, 2025",
    x = "Occupancy granted",
    y = "Gross assessed value uplift",
    colour = NULL,
    caption = CAPTION_COE
  ) +
  theme_jd(mode = "light")
```

The measurement cohort for this post is the matched 2024 permits with occupancy granted by September 30, 2025. Everything else, including those October-December completions, goes into the "not yet fully reflected" group.

```{r}
#| label: cohort-split
final_cohort <- matched_2024_all |>
  filter(
    !is.na(occupancy_granted_date),
    occupancy_granted_date <= occupancy_cutoff
  )

not_yet_complete_matched <- matched_2024_all |>
  anti_join(final_cohort, by = "permit_row_id")

n_final_cohort <- nrow(final_cohort)
```

The measurement cohort is `r n_final_cohort` properties. As an example, @tbl-spotlight traces two properties through the process, from their pre-redevelopment assessed value, through a mid-construction 2025 assessment, to their completed 2026 value. The Alberta Avenue example received occupancy in November 2025, yet its 2026 value looks fully assessed. This shows that the lag in @fig-assessment-lag is a trend, not a rule, which is why the conservative September cutoff is used for measurement.

```{r}
#| label: tbl-spotlight
#| tbl-cap: "Two properties traced from pre-redevelopment (2024) through mid-construction (2025) to a completed rowhome (2026)"
spotlight_accounts <- c("4042701", "2049286")

matched_2024_all |>
  filter(account_number %in% spotlight_accounts) |>
  mutate(
    neighbourhood = str_to_title(neighbourhood),
    mill_class_1 = str_to_title(mill_class_1)
  ) |>
  select(
    neighbourhood,
    units_added,
    value_2024,
    value_2025,
    value_2026,
    mill_class_1
  ) |>
  arrange(neighbourhood) |>
  gt() |>
  fmt_currency(columns = c(value_2024, value_2025, value_2026), decimals = 0) |>
  cols_label(
    neighbourhood = "Neighbourhood",
    units_added = "Units",
    value_2024 = "2024 (before)",
    value_2025 = "2025 (mid-construction)",
    value_2026 = "2026 (completed)",
    mill_class_1 = "2026 assessment class"
  ) |>
  tab_header(
    title = "Two RS Zone rowhomes, before and after",
    subtitle = "Assessed value by year"
  ) |>
  finalize_gt() |>
  opt_stylize_jd()
```

### The rest of the 2024 cohort

The `r n_final_cohort` properties above only cover 2024 permits whose completed buildings the 2026 assessment has caught up with. The remaining matched permits, `r nrow(not_yet_complete_matched)` of them, are either still under construction or finished too late in 2025 for the assessment to reflect. Their 2026 assessed values still tell us something: whatever value the construction had reached, a partial uplift. @tbl-2024-actual-total combines both groups into a single, fully-observed total for the 2024 permit year.

```{r}
#| label: tbl-2024-actual-total
#| tbl-cap: "Actual (not model-based) 2024 cohort assessed value uplift, split by completion status"
status_complete <- "Occupancy by Sept 2025"
status_partial <- "Occupancy after Sept 2025"

actual_2024 <- matched_2024_all |>
  mutate(
    status = if_else(
      permit_row_id %in% final_cohort$permit_row_id,
      status_complete,
      status_partial
    )
  ) |>
  select(status, gross_uplift)

actual_2024_summary <- actual_2024 |>
  mutate(status = fct_relevel(status, status_complete, status_partial)) |>
  group_by(status) |>
  summarise(
    n = n(),
    median_gross_uplift = median(gross_uplift),
    total_gross_uplift = sum(gross_uplift),
    .groups = "drop"
  ) |>
  arrange(status) |>
  mutate(status = as.character(status))

gt(actual_2024_summary, rowname_col = "status") |>
  grand_summary_rows(
    fns = "sum",
    fmt = list(
      ~ fmt_currency(
        .,
        columns = c(median_gross_uplift, total_gross_uplift),
        decimals = 0
      )
    )
  ) |>
  fmt_currency(
    columns = c(median_gross_uplift, total_gross_uplift),
    decimals = 0
  ) |>
  cols_label(
    n = "n",
    median_gross_uplift = "Median gross uplift",
    total_gross_uplift = "Total gross uplift"
  ) |>
  tab_header(
    title = "2024 cohort, actual gross uplift observed so far",
    subtitle = "No extrapolation model applied"
  ) |>
  finalize_gt() |>
  opt_stylize_jd()
```

That gives a fully-observed total of `r label_dollars_adaptive(sum(actual_2024_summary$total_gross_uplift))` in assessed value uplift already showing up in the 2026 assessment from the 2024 cohort alone. It understates the eventual value of that cohort: the "not yet fully reflected" group is likely still a stale, mid-construction assessment rather than the finished-building value — exactly the gap the model-based extrapolation later in this post is designed to estimate.

### From gross uplift to a fairer comparison

The raw 2024-to-2026 difference in assessed value overstates the effect of redevelopment on its own, because Edmonton property values were also generally rising over that period. To isolate the redevelopment effect, I build a comparison: what would each property have been worth in 2026 if it hadn't been redeveloped? I estimate this from a group of RS-zoned properties in mature neighbourhoods that were not part of any 2024-2026 residential new-build permit, taking the median ratio of their 2026 to 2024 assessed value as a general appreciation factor. This un-redeveloped control group may not appreciate at exactly the rate the redeveloped lots would have, but it's a reasonable counterfactual, and good enough for what I'm trying to measure here.

```{r}
#| label: counterfactual-growth
redeveloped_legal_keys <- bp |>
  st_drop_geometry() |>
  filter(units_added >= 1, work_type == "(01) Building - New") |>
  distinct(legal_description) |>
  pull(legal_description)

nbhd_type_lookup <- bp |>
  st_drop_geometry() |>
  distinct(neighbourhood, neighbourhood_type) |>
  filter(!is.na(neighbourhood_type))

control_group <- assessment_2024 |>
  filter(zoning == "RS") |>
  left_join(nbhd_type_lookup, by = c("neighbourhood_name" = "neighbourhood")) |>
  filter(
    neighbourhood_type == "Mature",
    !legal_description %in% redeveloped_legal_keys,
    !account_number %in% final_cohort$account_number
  ) |>
  inner_join(
    assessment_2026 |> select(account_number, value_2026 = assessed_value),
    by = "account_number"
  ) |>
  mutate(growth_ratio = value_2026 / assessed_value)

g <- median(control_group$growth_ratio, na.rm = TRUE)

final_cohort <- final_cohort |>
  mutate(
    gross_uplift = value_2026 - value_2024,
    net_uplift = value_2026 - g * value_2024
  )
```

Across `r comma(nrow(control_group))` non-redeveloped RS Zone properties in mature neighbourhoods, the median 2024-to-2026 growth ratio is `r round(g, 2)`x, meaning a typical mature-neighbourhood house that wasn't touched still gained about `r percent(g - 1, accuracy = 1)` in assessed value over those two years. I use this as the baseline against which redevelopment uplift is measured. Net uplift is the completed rowhome's 2026 value minus `r round(g, 2)` times its 2024 baseline, isolating the redevelopment effect from general market appreciation. Gross uplift is the simple, unadjusted difference, shown alongside it for comparison.

@fig-uplift-distribution shows the distribution of both measures across the `r n_final_cohort` matched properties, split by project size.

```{r}
#| label: fig-uplift-distribution
#| fig-cap: "Distribution of assessed value uplift per redeveloped property, gross and net of market appreciation"
uplift_long <- final_cohort |>
  select(bucket, gross_uplift, net_uplift) |>
  pivot_longer(
    c(gross_uplift, net_uplift),
    names_to = "type",
    values_to = "uplift"
  ) |>
  mutate(
    type = if_else(
      type == "gross_uplift",
      "Gross uplift",
      "Net of appreciation"
    )
  )

ggplot(uplift_long, aes(x = bucket, y = uplift, fill = type)) +
  geom_boxplot(outlier.alpha = 0.4, position = position_dodge(width = 0.75)) +
  scale_y_continuous(labels = label_dollar(scale = 1e-6, suffix = "M")) +
  labs(
    title = "Assessed value uplift per redeveloped property",
    subtitle = "2024 RS Zone 5-8 unit rowhomes, mature neighbourhoods,\noccupancy by Sept 2025",
    x = NULL,
    y = "Assessed value uplift",
    fill = NULL,
    caption = CAPTION_COE
  ) +
  theme_jd(mode = "light")
```

@tbl-uplift-rules compares three simple ways of summarizing this uplift into a single per-project figure: a median per property within two size buckets (5-6 units and 7-8 units), a single median per property across all sizes, and a median per home added, which can be scaled by any project's unit count. These serve as sanity anchors for the model-based rule developed next.

```{r}
#| label: tbl-uplift-rules
#| tbl-cap: "Three ways to summarize per-project uplift, gross and net of appreciation"
rule_bucketed <- final_cohort |>
  group_by(bucket) |>
  summarise(
    n = n(),
    gross = median(gross_uplift),
    net = median(net_uplift),
    .groups = "drop"
  ) |>
  rename(rule = bucket)

rule_pooled <- final_cohort |>
  summarise(n = n(), gross = median(gross_uplift), net = median(net_uplift)) |>
  mutate(rule = "All together", .before = 1)

rule_per_unit <- final_cohort |>
  summarise(
    n = n(),
    gross = median(gross_uplift / units_added),
    net = median(net_uplift / units_added)
  ) |>
  mutate(rule = "Per home added", .before = 1)

bind_rows(rule_bucketed, rule_pooled, rule_per_unit) |>
  gt(rowname_col = "rule") |>
  fmt_currency(columns = c(gross, net), decimals = 0) |>
  cols_label(
    n = "n",
    gross = "Median gross uplift",
    net = "Median net uplift"
  ) |>
  tab_header(
    title = "Three uplift rules compared",
    subtitle = "Measured from the completed 2024 cohort"
  ) |>
  finalize_gt() |>
  opt_stylize_jd()
```

### A model for uplift {#sec-model-for-uplift}

The simple rules above ignore everything about a project except its unit count. But uplift plausibly varies with location too: land near LRT stations and in more expensive neighbourhoods could command different completed values. So alongside the flat medians, I fit a simple regression of each property's gross uplift on three things: units added, distance from the nearest LRT stop, and the neighbourhood's median RS-zone assessed value (a land-value measure computed from the same 2024 assessment data).

This isn't an attempt at a precise or causal model of what drives redevelopment value — with only `r n_final_cohort` observations, it can't be, and that's not the point. It's a deliberately simple, "good enough" prediction rule, and the standard it needs to clear is agreement with the flat per-project medians above, with the payoff of a little more precision: a prediction tailored to each permit's own size and location rather than one number for every project of a given size. [Specification details — which covariates and transforms were tried, and how this model compares against simpler and more flexible alternatives by cross-validation — are in the @sec-model-appendix.]{.aside}

```{r}
#| label: uplift-model
# Neighbourhood land-value covariate: median 2024 assessed value of
# RS-zoned properties, per neighbourhood.
nbhd_median_rs_value <- assessment_2024 |>
  filter(zoning == "RS") |>
  group_by(neighbourhood_name) |>
  summarise(nbhd_median_value = median(assessed_value), .groups = "drop")

final_cohort <- final_cohort |>
  left_join(
    nbhd_median_rs_value,
    by = c("neighbourhood" = "neighbourhood_name")
  )

stopifnot(
  all(!is.na(final_cohort$nbhd_median_value)),
  all(!is.na(final_cohort$distance_from_lrt)),
  all(final_cohort$gross_uplift > 0)
)

loocv_rmse <- function(fit_fun, pred_fun, data) {
  errs <- vapply(
    seq_len(nrow(data)),
    function(i) {
      m <- fit_fun(data[-i, ])
      pred_fun(m, data[i, ]) - data$gross_uplift[i]
    },
    numeric(1)
  )
  sqrt(mean(errs^2))
}
# Same as loocv_rmse() but returns the per-property errors rather than their
# RMSE, needed for a paired bootstrap comparison between two specs.
loocv_errs <- function(fit_fun, pred_fun, data) {
  vapply(
    seq_len(nrow(data)),
    function(i) {
      m <- fit_fun(data[-i, ])
      pred_fun(m, data[i, ]) - data$gross_uplift[i]
    },
    numeric(1)
  )
}

# Covariate-transform search: units added and neighbourhood value each on a
# linear vs. log scale, distance from LRT on a linear vs. inverse-distance
# scale, always with gross uplift itself left untransformed (see prose
# below). errs_level/errs_level_log are kept per-observation to support a
# paired bootstrap comparison between the previous default and the winner.
errs_level <- loocv_errs(
  \(d) {
    lm(
      gross_uplift ~ units_added +
        distance_from_lrt +
        I(nbhd_median_value / 1e5),
      d
    )
  },
  \(m, nd) predict(m, nd),
  final_cohort
)
errs_level_log <- loocv_errs(
  \(d) {
    lm(
      gross_uplift ~ log(units_added) +
        distance_from_lrt +
        log(nbhd_median_value),
      d
    )
  },
  \(m, nd) predict(m, nd),
  final_cohort
)
rmse_level <- sqrt(mean(errs_level^2))
rmse_level_log <- sqrt(mean(errs_level_log^2))
rmse_log <- loocv_rmse(
  \(d) {
    lm(
      log(gross_uplift) ~ units_added +
        distance_from_lrt +
        log(nbhd_median_value),
      d
    )
  },
  # Duan smearing estimator for the back-transform
  \(m, nd) exp(predict(m, nd)) * mean(exp(residuals(m))),
  final_cohort
)
rmse_median <- sqrt(mean(
  (median(final_cohort$gross_uplift) - final_cohort$gross_uplift)^2
))

# Paired bootstrap SE of an RMSE difference (dollars, matching rmse_*'s own
# units) between two specs fit on the same 48 properties.
paired_bootstrap_se <- function(errs_a, errs_b, B = 2000, seed = 1) {
  n <- length(errs_a)
  old_seed <- .Random.seed
  set.seed(seed)
  diffs <- replicate(B, {
    idx <- sample.int(n, n, replace = TRUE)
    sqrt(mean(errs_a[idx]^2)) - sqrt(mean(errs_b[idx]^2))
  })
  .Random.seed <<- old_seed
  sd(diffs)
}
spec_choice_se <- paired_bootstrap_se(errs_level_log, errs_level)

uplift_model <- lm(
  gross_uplift ~ log(units_added) + distance_from_lrt + log(nbhd_median_value),
  data = final_cohort
)
```

As a sanity check against @tbl-uplift-rules: the model's average predicted uplift across the measured cohort is `r label_dollars_adaptive(mean(fitted(uplift_model)))` per property, right in line with the simple medians, as it should be. What the model adds is the ability to give each individual permit a prediction tailored to its size and location, which matters when extrapolating to hundreds of permits whose size mix and geography differ from the measured cohort's.

For net uplift, the counterfactual adjustment is applied after prediction: a property's predicted net uplift is its predicted gross uplift minus (`r round(g, 2)` − 1) times its own 2024 baseline value, the same adjustment used for the measured cohort.

## From one cohort to the whole pipeline

The measured cohort only covers 2024 permits that were both matchable and fully reflected in the 2026 assessment. To estimate the cumulative effect of the RS Zone rowhome since ZBR took effect, I apply the uplift model (coefficients in @tbl-uplift-model, appendix) to every other mature-neighbourhood RS Zone 5-8 unit new-build permit issued since 2024: the rest of the 2024 cohort (permits that didn't match or aren't fully assessed yet), all of 2025, and 2026. Each permit gets a prediction from its own unit count, LRT distance, and neighbourhood.

This is a **steady-state estimate**: it assumes every one of those permits eventually reaches the completed-building uplift the model predicts, once construction wraps up and a future assessment catches up to it. It is not the smaller amount those properties currently show in the 2026 roll, which @tbl-2024-actual-total showed directly for 2024's not-yet-reflected permits. The gap between that actual, partial total and the steady-state estimate below is construction and assessment lag, not double-counting.

```{r}
#| label: pipeline-counts
pipeline_counts <- rowhome_pipeline |>
  st_drop_geometry() |>
  mutate(bucket = if_else(units_added <= 6, "5-6 units", "7-8 units")) |>
  count(year, bucket)

# H1 annualization multiplier, same method as the Q2 2026 report: median
# ratio of full-year to H1 residential units across historical years.
h1_ratios <- bp |>
  st_drop_geometry() |>
  filter(year < 2026) |>
  group_by(year) |>
  summarize(
    full_year = sum(units_added),
    h1 = sum(units_added[month_number <= 6]),
    .groups = "drop"
  ) |>
  mutate(ratio = full_year / h1)
h1_multiplier <- round(median(h1_ratios$ratio), 1)

n_2024_all <- pipeline_counts |>
  filter(year == 2024) |>
  summarise(n = sum(n)) |>
  pull(n)
n_2025_all <- pipeline_counts |>
  filter(year == 2025) |>
  summarise(n = sum(n)) |>
  pull(n)
n_2026_h1 <- pipeline_counts |>
  filter(year == 2026) |>
  summarise(n = sum(n)) |>
  pull(n)
```

2025 permitted **`r n_2025_all`** more mature-neighbourhood RS Zone 5-8 unit rowhomes, and H1 2026 has already permitted `r n_2026_h1`. Following the same [H1 annualization method used in the Q2 2026 report](../q2-building-permits/index.qmd), full-year 2026 is projected at `r h1_multiplier`x the H1 count.

Predicting net uplift needs each permit's own 2024 baseline value, so the 2025 and 2026 permits are run through the same baseline matchers as the 2024 cohort (their lots all existed in the 2024 roll, pre-redevelopment).

```{r}
#| label: extrapolation
# Blended municipal tax rate per size bucket, weighted by each bucket's
# observed mix of Residential vs. Other Residential assessment class. Used
# both for the extrapolated tax figures below and, via bucket_rates_2026,
# for the neighbourhood breakdown later in the post.
bucket_rates_2026 <- final_cohort |>
  mutate(bucket = if_else(units_added <= 6, "5-6 units", "7-8 units")) |>
  count(bucket, mill_class_1) |>
  group_by(bucket) |>
  mutate(share = n / sum(n)) |>
  summarise(
    rate = sum(
      share *
        if_else(
          mill_class_1 == "OTHER RESIDENTIAL",
          municipal_rate("Other Residential", 2026),
          municipal_rate("Residential", 2026)
        )
    ),
    .groups = "drop"
  )

bucket_rates_2025 <- final_cohort |>
  mutate(bucket = if_else(units_added <= 6, "5-6 units", "7-8 units")) |>
  count(bucket, mill_class_1) |>
  group_by(bucket) |>
  mutate(share = n / sum(n)) |>
  summarise(
    rate = sum(
      share *
        if_else(
          mill_class_1 == "OTHER RESIDENTIAL",
          municipal_rate("Other Residential", 2025),
          municipal_rate("Residential", 2025)
        )
    ),
    .groups = "drop"
  )

measured_ids <- final_cohort$permit_row_id

# Baselines for 2025/2026 permits, using the same matchers as the 2024 cohort
permits_2025_2026 <- rowhome_pipeline |> filter(year %in% 2025:2026)
future_baselines <- match_baseline_2024(permits_2025_2026) |>
  geo_fallback_baseline(permits_2025_2026) |>
  select(permit_row_id, value_2024)

baseline_match_rate <- mean(!is.na(future_baselines$value_2024))

# Every mature RS 5-8 unit new-build permit with the model's covariates.
# Fallbacks for prediction: median LRT distance for permits with no mapped
# location; the neighbourhood median RS value doubles as the baseline proxy
# for permits that didn't match one.
median_pipeline_distance <- median(
  rowhome_pipeline$distance_from_lrt,
  na.rm = TRUE
)

all_permits <- rowhome_pipeline |>
  st_drop_geometry() |>
  filter(year %in% 2024:2026) |>
  left_join(
    nbhd_median_rs_value,
    by = c("neighbourhood" = "neighbourhood_name")
  ) |>
  left_join(
    bind_rows(
      matched_2024_all |> select(permit_row_id, value_2024),
      future_baselines
    ),
    by = "permit_row_id"
  ) |>
  mutate(
    nbhd_median_value = coalesce(
      nbhd_median_value,
      median(nbhd_median_rs_value$nbhd_median_value)
    ),
    distance_from_lrt = coalesce(distance_from_lrt, median_pipeline_distance),
    baseline_value = coalesce(value_2024, nbhd_median_value),
    pred_gross = predict(
      uplift_model,
      newdata = pick(units_added, distance_from_lrt, nbhd_median_value)
    ),
    pred_net = pred_gross - (g - 1) * baseline_value,
    bucket = if_else(units_added <= 6, "5-6 units", "7-8 units")
  ) |>
  left_join(bucket_rates_2026, by = "bucket") |>
  rename(rate_2026 = rate) |>
  left_join(bucket_rates_2025, by = "bucket") |>
  rename(rate_2025 = rate) |>
  mutate(
    pred_tax_gross_2026 = pred_gross * rate_2026,
    pred_tax_net_2026 = pred_net * rate_2026,
    pred_tax_net_2025 = pred_net * rate_2025
  )

measured_totals <- final_cohort |>
  summarise(gross = sum(gross_uplift), net = sum(net_uplift))

model_based_2024 <- all_permits |>
  filter(year == 2024, !permit_row_id %in% measured_ids) |>
  summarise(
    gross = sum(pred_gross),
    net = sum(pred_net),
    tax_gross_2026 = sum(pred_tax_gross_2026),
    tax_net_2026 = sum(pred_tax_net_2026),
    tax_net_2025 = sum(pred_tax_net_2025)
  )

model_based_2025 <- all_permits |>
  filter(year == 2025) |>
  summarise(
    gross = sum(pred_gross),
    net = sum(pred_net),
    tax_gross_2026 = sum(pred_tax_gross_2026),
    tax_net_2026 = sum(pred_tax_net_2026),
    tax_net_2025 = sum(pred_tax_net_2025)
  )

model_based_2026_h1 <- all_permits |>
  filter(year == 2026) |>
  summarise(
    gross = sum(pred_gross),
    net = sum(pred_net),
    tax_gross_2026 = sum(pred_tax_gross_2026),
    tax_net_2026 = sum(pred_tax_net_2026),
    tax_net_2025 = sum(pred_tax_net_2025)
  )

# 2026 full-year projection: no per-permit rows exist yet for the
# un-permitted remainder of the year, so scale H1's predicted total.
model_based_2026_projected_remainder <- model_based_2026_h1 |>
  mutate(
    gross = gross * (h1_multiplier - 1),
    net = net * (h1_multiplier - 1),
    tax_gross_2026 = tax_gross_2026 * (h1_multiplier - 1),
    tax_net_2026 = tax_net_2026 * (h1_multiplier - 1)
  )

total_gross_uplift <- measured_totals$gross +
  model_based_2024$gross +
  model_based_2025$gross +
  model_based_2026_h1$gross +
  model_based_2026_projected_remainder$gross
total_net_uplift <- measured_totals$net +
  model_based_2024$net +
  model_based_2025$net +
  model_based_2026_h1$net +
  model_based_2026_projected_remainder$net

n_2026_projected <- n_2026_h1 * h1_multiplier
```

The 2025 and 2026 permits match their 2024 baselines at a **`r percent(baseline_match_rate, accuracy = 1)`** rate; the few that don't use their neighbourhood's median RS value as a stand-in baseline for the net adjustment.

@fig-cumulative-value-uplift shows the resulting cumulative assessed value uplift by permit year, split between the actually-measured portion and the portion estimated with the model.

```{r}
#| label: fig-cumulative-value-uplift
#| fig-cap: "Cumulative estimated assessed value uplift since ZBR, by permit year (net of appreciation)"
#| classes: "preview-image"
cumulative_by_year <- tibble(
  year = c(2024, 2024, 2025, 2026, 2026),
  source = c(
    "Measured",
    "Estimated (model)",
    "Estimated (model)",
    "Estimated (model)",
    "Estimated (model)"
  ),
  net_uplift = c(
    measured_totals$net,
    model_based_2024$net,
    model_based_2025$net,
    model_based_2026_h1$net,
    model_based_2026_projected_remainder$net
  )
) |>
  group_by(year, source) |>
  summarise(net_uplift = sum(net_uplift), .groups = "drop")

ggplot(
  cumulative_by_year,
  aes(x = factor(year), y = net_uplift, fill = source)
) +
  geom_col() +
  scale_y_continuous(labels = label_dollar(scale = 1e-6, suffix = "M")) +
  labs(
    title = "Cumulative assessed value uplift since ZBR",
    subtitle = paste0(
      "Net of market appreciation; 2026 full year projected at ",
      h1_multiplier,
      "x H1"
    ),
    x = "Permit year",
    y = "Assessed value uplift",
    fill = NULL,
    caption = CAPTION_COE
  ) +
  theme_jd(mode = "light")
```

Adding it all up: since ZBR took effect, mature-neighbourhood RS Zone 5-8 unit rowhomes are estimated to have added **`r label_dollars_adaptive(total_gross_uplift)`** in gross assessed value, or **`r label_dollars_adaptive(total_net_uplift)`** net of general market appreciation, across `r n_2024_all` (2024) + `r n_2025_all` (2025) + `r round(n_2026_projected)` (2026, projected) = `r n_2024_all + n_2025_all + round(n_2026_projected)` permitted projects.

```{r}
#| label: cumulative-by-neighbourhood-prep
#| include: false
# Each actual permit's uplift: measured where available, model prediction
# otherwise. The 2026 projected remainder has no real permits to attach
# to a neighbourhood, so it's excluded here (it's still in the citywide
# total above). Used by the neighbourhood breakdown in "Where is this
# happening?" below.
measured_lookup <- final_cohort |>
  st_drop_geometry() |>
  select(
    permit_row_id,
    actual_gross = gross_uplift,
    actual_net = net_uplift,
    mill_class_1,
    value_2024,
    value_2026
  )

permit_uplift <- all_permits |>
  select(-value_2024) |>
  left_join(measured_lookup, by = "permit_row_id") |>
  mutate(
    is_measured = !is.na(actual_gross),
    gross_uplift = coalesce(actual_gross, pred_gross),
    net_uplift = coalesce(actual_net, pred_net),
    annual_tax_uplift = if_else(
      is_measured,
      if_else(
        mill_class_1 == "OTHER RESIDENTIAL",
        municipal_rate("Other Residential", 2026) * value_2026,
        municipal_rate("Residential", 2026) * value_2026
      ) -
        municipal_rate("Residential", 2026) * g * value_2024,
      pred_tax_net_2026
    )
  )

cumulative_by_neighbourhood <- permit_uplift |>
  mutate(neighbourhood = str_to_title(neighbourhood)) |>
  group_by(neighbourhood) |>
  summarise(
    n_projects = n(),
    homes_added = sum(units_added),
    cumulative_gross_uplift = sum(gross_uplift),
    cumulative_net_uplift = sum(net_uplift),
    annual_tax_uplift = sum(annual_tax_uplift),
    .groups = "drop"
  )
```

## What this means for property tax

Assessed value only matters for the City's budget insofar as it's taxed. I convert the estimated uplift into two figures: an **annual recurring** estimate, at 2026 municipal tax rates, of how much more property tax this pipeline will generate every year once every project is complete and fully reflected in an assessment roll; and a rougher **cumulative collected so far** estimate that accounts for the fact that most of this pipeline is still under construction or only recently completed.

Completed 5-8 unit rowhomes on a single title are typically assessed as "Other Residential", which carries a higher municipal mill rate than the "Residential" class the original house was assessed under, but that split isn't uniform: in the measured cohort, essentially every 7-8 unit building lands in Other Residential, while most 5-6 unit buildings stay Residential. In the measured cohort overall, `r percent(final_cohort |> summarise(p = mean(mill_class_1 == "OTHER RESIDENTIAL")) |> pull(p), accuracy = 1)` of completed properties land in Other Residential; for the model-based portion of the estimate, I apply that same mix separately within the 5-6 and 7-8 unit buckets, rather than one flat rate for every permit regardless of size.

```{r}
#| label: tax-conversion
tax_measured <- final_cohort |>
  mutate(
    new_rate = if_else(
      mill_class_1 == "OTHER RESIDENTIAL",
      municipal_rate("Other Residential", 2026),
      municipal_rate("Residential", 2026)
    ),
    annual_tax_gross = new_rate *
      value_2026 -
      municipal_rate("Residential", 2026) * value_2024,
    annual_tax_net = new_rate *
      value_2026 -
      municipal_rate("Residential", 2026) * g * value_2024
  ) |>
  summarise(gross = sum(annual_tax_gross), net = sum(annual_tax_net))

tax_model_based_gross <- model_based_2024$tax_gross_2026 +
  model_based_2025$tax_gross_2026 +
  model_based_2026_h1$tax_gross_2026 +
  model_based_2026_projected_remainder$tax_gross_2026
tax_model_based_net <- model_based_2024$tax_net_2026 +
  model_based_2025$tax_net_2026 +
  model_based_2026_h1$tax_net_2026 +
  model_based_2026_projected_remainder$tax_net_2026

annual_tax_uplift_gross <- tax_measured$gross + tax_model_based_gross
annual_tax_uplift_net <- tax_measured$net + tax_model_based_net
```

At 2026 municipal tax rates, this pipeline is estimated to add **`r label_dollars_adaptive(annual_tax_uplift_net)`** in property tax every year, net of what general appreciation would have generated anyway (`r label_dollars_adaptive(annual_tax_uplift_gross)` on a gross basis). That's before counting the provincial education levy, which would add roughly another third on top of the municipal portion.

Because most of this pipeline is still ramping up, actual tax collected so far is much lower than that steady-state figure. Comparing each matched property's mid-construction 2025 assessed value to its 2024 baseline and completed 2026 value shows how much of the total uplift typically shows up one year after a permit is issued.

```{r}
#| label: cumulative-collected
partial_uplift_share <- final_cohort |>
  filter(!is.na(value_2025), value_2026 > value_2024) |>
  mutate(p = (value_2025 - value_2024) / (value_2026 - value_2024)) |>
  summarise(p = median(p)) |>
  pull(p) |>
  pmax(0) |>
  pmin(1)

# Measured properties use their own exact assessment class; model-based
# permits use the bucketed blended rate (see all_permits' pred_tax_net_2025/
# pred_tax_net_2026 columns).
measured_tax_2025 <- final_cohort |>
  summarise(
    tax = sum(
      net_uplift *
        if_else(
          mill_class_1 == "OTHER RESIDENTIAL",
          municipal_rate("Other Residential", 2025),
          municipal_rate("Residential", 2025)
        )
    )
  ) |>
  pull(tax)

measured_tax_2026 <- final_cohort |>
  summarise(
    tax = sum(
      net_uplift *
        if_else(
          mill_class_1 == "OTHER RESIDENTIAL",
          municipal_rate("Other Residential", 2026),
          municipal_rate("Residential", 2026)
        )
    )
  ) |>
  pull(tax)

# A property permitted in year Y contributes a partial share of its
# annual tax uplift in year Y+1, and the full amount from Y+2 on.
collected_2025 <- partial_uplift_share *
  (measured_tax_2025 + model_based_2024$tax_net_2025)
collected_2026 <- (measured_tax_2026 + model_based_2024$tax_net_2026) +
  partial_uplift_share * model_based_2025$tax_net_2026
cumulative_collected <- collected_2025 + collected_2026
```

The median property has captured about `r percent(partial_uplift_share, accuracy = 1)` of its full uplift by the assessment one year after permitting, consistent with most projects still being under construction at that point. Modelling each year's cohort as contributing that partial share in the following year and the full amount from two years out, I estimate Edmonton has collected roughly **`r label_dollars_adaptive(cumulative_collected)`** in incremental property tax from this pipeline across 2025 and 2026 combined, well short of the `r label_dollars_adaptive(annual_tax_uplift_net)`-a-year steady-state run rate established above.

These are estimates, not City budget figures. Assessment growth from new construction doesn't translate one-for-one into new City revenue, since Council sets the mill rate each year based on total budget needs, but new construction is genuinely additional tax base at whatever rate Council sets, distinct from a rate increase on existing properties. This also doesn't account for the cost of servicing new residents, which isn't the subject of this post.

## Where is this happening?

That citywide total is the sum of hundreds of individual redevelopments, concentrated unevenly across mature Edmonton. @tbl-cumulative-selected-neighbourhoods breaks the cumulative estimate down for the same neighbourhoods used in the [Q2 2026 report's selected-neighbourhood table](../q2-building-permits/index.qmd), places that have been prominent in debates about RS Zone infill. Each permit is valued at its actual measured uplift where available, or the model's prediction otherwise, then summed by neighbourhood. This covers every actual permit issued through H1 2026; the projected remainder of 2026 isn't tied to real permits yet, so it's included in the citywide total above but left out of this neighbourhood breakdown.

```{r}
#| label: tbl-cumulative-selected-neighbourhoods
#| tbl-cap: "Cumulative estimated uplift for selected neighbourhoods, 2024 through H1 2026"
selected_nbhds <- c(
  "GLENWOOD",
  "GROVENOR",
  "BRITANNIA YOUNGSTOWN",
  "WESTMOUNT",
  "MCKERNAN",
  "BELGRAVIA",
  "CRESTWOOD",
  "PARKVIEW"
)

tibble(neighbourhood = str_to_title(selected_nbhds)) |>
  left_join(cumulative_by_neighbourhood, by = "neighbourhood") |>
  replace_na(list(
    n_projects = 0,
    homes_added = 0,
    cumulative_gross_uplift = 0,
    cumulative_net_uplift = 0,
    annual_tax_uplift = 0
  )) |>
  arrange(neighbourhood) |>
  gt() |>
  fmt_currency(
    columns = c(
      cumulative_gross_uplift,
      cumulative_net_uplift,
      annual_tax_uplift
    ),
    decimals = 0
  ) |>
  fmt_integer(columns = c(n_projects, homes_added)) |>
  cols_label(
    neighbourhood = "Neighbourhood",
    n_projects = "Projects",
    homes_added = "Homes added",
    cumulative_gross_uplift = "Cumulative gross uplift",
    cumulative_net_uplift = "Cumulative net uplift",
    annual_tax_uplift = "Est. annual municipal tax uplift"
  ) |>
  tab_header(
    title = "Cumulative uplift, selected neighbourhoods",
    subtitle = "2024 through H1 2026, measured + model-based"
  ) |>
  finalize_gt() |>
  opt_stylize_jd() |>
  sub_missing()
```

@tbl-uplift-by-neighbourhood is the same breakdown for every mature neighbourhood with a matching permit, not just the selected set above; use it to look up any specific neighbourhood.

```{r}
#| label: tbl-uplift-by-neighbourhood
#| tbl-cap: "Cumulative estimated uplift by neighbourhood, 2024 through H1 2026 (interactive)"
cumulative_by_neighbourhood |>
  arrange(desc(annual_tax_uplift)) |>
  gt() |>
  fmt_currency(
    columns = c(
      cumulative_gross_uplift,
      cumulative_net_uplift,
      annual_tax_uplift
    ),
    decimals = 0
  ) |>
  fmt_integer(columns = c(n_projects, homes_added)) |>
  cols_label(
    neighbourhood = "Neighbourhood",
    n_projects = "Projects",
    homes_added = "Homes added",
    cumulative_gross_uplift = "Cumulative gross uplift",
    cumulative_net_uplift = "Cumulative net uplift",
    annual_tax_uplift = "Est. annual municipal tax uplift"
  ) |>
  tab_header(
    title = "Cumulative uplift by neighbourhood",
    subtitle = "2024 through H1 2026, measured + model-based"
  ) |>
  finalize_gt(interactive = TRUE) |>
  opt_stylize_jd()
```

```{r}
#| label: key-facts-validation
#| include: false
# Sanity check that the hardcoded Key facts bullets above still match the
# computed figures, so a future re-render can't silently drift from them.
n_total_projects <- n_2024_all + n_2025_all + round(n_2026_projected)
stopifnot(
  total_net_uplift > 1.2e9,
  total_net_uplift < 1.4e9,
  annual_tax_uplift_net > 9.5e6,
  annual_tax_uplift_net < 11.5e6,
  cumulative_collected > 1.5e6,
  cumulative_collected < 2.5e6,
  n_total_projects > 750,
  n_total_projects < 850
)
```

## Conclusions

The RS Zone rowhome isn't just adding homes, it's rebuilding the property tax base of neighbourhoods that had been losing population and, with it, tax revenue per hectare of city-maintained infrastructure for decades. A handful of redevelopments per neighbourhood, concentrated in mature areas that already have the pipes, roads, and schools in place, is already worth more than `r label_dollars_adaptive(annual_tax_uplift_net)` a year in additional municipal property tax capacity at steady state, on top of the housing it delivers.

None of this required a subsidy, a grant, or a City-led development. It required removing rules that made this kind of building illegal. If you'd like to help keep it that way, consider joining [Grow Together Edmonton](https://www.growtogetheryeg.com/).

## Appendix: model specification details {#sec-model-appendix}

The regression used to extrapolate uplift across the pipeline ([above](#sec-model-for-uplift)) is a deliberately simple prediction rule, not a causal model — its job is to agree with the flat per-project medians while adding a bit of size- and location-specific precision. This appendix documents the two specification choices behind it, for readers who want the detail.

First, modelling gross uplift is algebraically the same as modelling the completed 2026 value with the 2024 baseline as a fixed offset. Letting the baseline enter with a freely estimated coefficient instead doesn't help: the old house's value is nearly collinear with the neighbourhood land-value measure (r = `r round(cor(final_cohort$value_2024, final_cohort$nbhd_median_value), 2)`), so the offset form is both more stable and better in cross-validation.

Second, the uplift itself is *not* log-transformed: the measured cohort's uplift distribution isn't right-skewed enough to justify it, and a log-log variant does worse in cross-validation (see @tbl-model-comparison). But its two *covariates* are worth transforming: I searched units added and neighbourhood value on both linear and log scales, and distance from LRT on both linear and inverse-distance scales, always keeping gross uplift itself on the dollar scale, evaluated by leave-one-out cross-validation on the dollar scale. @tbl-model-comparison shows the two ends of that search. Log-transforming units added and neighbourhood value together gives the best cross-validated fit, an improvement over the plain linear specification and clearly ahead of log-transforming the outcome instead. With only `r n_final_cohort` observations, that improvement is real but not dramatic — about `r round((rmse_level - rmse_level_log) / spec_choice_se, 1)` times its own bootstrap standard error, so I adopt it as the best available specification on the evidence, not because the gap is large.

```{r}
#| label: tbl-model-comparison
#| tbl-cap: "Comparing uplift model specifications by leave-one-out cross-validated RMSE"
tibble(
  spec = c(
    "Units + LRT distance + neighbourhood value (linear)",
    "log(units) + LRT distance + log(neighbourhood value) (adopted)",
    "log(gross uplift) ~ units + LRT distance + log(neighbourhood value)",
    "Pooled median (no covariates)"
  ),
  rmse = c(rmse_level, rmse_level_log, rmse_log, rmse_median)
) |>
  gt(rowname_col = "spec") |>
  fmt_currency(columns = rmse, decimals = 0) |>
  cols_label(rmse = "LOOCV RMSE") |>
  tab_header(
    title = "Comparing uplift model specifications",
    subtitle = paste0(
      "Leave-one-out cross-validated RMSE, dollar scale, n = ",
      n_final_cohort
    )
  ) |>
  finalize_gt() |>
  opt_stylize_jd()
```

@tbl-uplift-model shows the chosen model's coefficients.

```{r}
#| label: tbl-uplift-model
#| tbl-cap: "Gross uplift model coefficients, fit on the measured 2024 cohort"
broom_coefs <- summary(uplift_model)$coefficients |>
  as_tibble(rownames = "term") |>
  rename(
    estimate = Estimate,
    se = `Std. Error`,
    t = `t value`,
    p = `Pr(>|t|)`
  ) |>
  mutate(
    term = case_match(
      term,
      "(Intercept)" ~ "Intercept",
      "log(units_added)" ~ "Units added (log)",
      "distance_from_lrt" ~ "Distance from LRT (per km)",
      "log(nbhd_median_value)" ~ "Neighbourhood median RS value (log)"
    )
  )

broom_coefs |>
  gt(rowname_col = "term") |>
  fmt_currency(columns = c(estimate, se), decimals = 0) |>
  fmt_number(columns = t, decimals = 2) |>
  fmt_number(columns = p, decimals = 3) |>
  cols_label(
    estimate = "Estimate",
    se = "Std. error",
    t = "t",
    p = "p-value"
  ) |>
  tab_header(
    title = "What drives per-property uplift?",
    subtitle = paste0(
      "Gross uplift ~ log(units) + LRT distance + log(neighbourhood value); R² = ",
      round(summary(uplift_model)$r.squared, 2),
      ", n = ",
      n_final_cohort
    )
  ) |>
  finalize_gt() |>
  opt_stylize_jd()
```